---
title: "Supplemental figures"
author: "LJ"
date: "2026-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Figure S1

## Bacteria

## Panel A - Bacteria

```{r}
###########################################################################################
# transform the data
ps_B_tr <- microbiome::transform(physeq_B, 'hellinger')

# PCoA - with Bray Curtis
pcoa_B_tr <- ordinate(ps_B_tr, "PCoA", distance = "bray", binary = FALSE)

# Extract PCoA results
pcoa_result_B_tr <- as.data.frame(pcoa_B_tr$vectors)

# Extract sample data for temporal information
temporal_data_B_tr <- as.data.frame(sample_data(ps_B_tr))

# Combine
combined_data_B_tr <- cbind(temporal_data_B_tr, pcoa_result_B_tr)

# Centroids
centroid_data_B_tr <- combined_data_B_tr %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Factor order for time
combined_data_B_tr$Timepoint <- factor(combined_data_B_tr$Timepoint,
                                       levels = c("T0","T1","T2","T3"))

arranged_data_B_tr <- centroid_data_B_tr %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_B_tr)$bulk_rhizo_t_2 <- paste0(sample_data(ps_B_tr)$bulk_rhizo, "_",
                                              sample_data(ps_B_tr)$Timepoint)

# Panel A

S1_B_A <- ggplot(combined_data_B_tr, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo_t_2), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_B_tr$values$Eigenvalues[1] / sum(pcoa_B_tr$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_B_tr$values$Eigenvalues[2] / sum(pcoa_B_tr$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

ggsave(paste0(workdir, "plots/pcoa/S1_B_PanelA_", Sys.Date(), ".pdf"),
       device = "pdf", plot = S1_B_A, width = 7, height = 6, dpi = 600)

```


## Panel B

```{r}

# PCoA - with Bray Curtis

### T1
ps_B_tr_T1 <- ps_filter(ps_B_tr, Timepoint == "T1")

pcoa_B_tr_T1 <- ordinate(ps_B_tr_T1, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_B_tr_T1 <- as.data.frame(pcoa_B_tr_T1$vectors)

# Extract sample data for temporal information
temporal_data_B_tr_T1 <- as.data.frame(sample_data(ps_B_tr_T1))

# Combine data for plotting
combined_data_B_tr_T1 <- cbind(temporal_data_B_tr_T1, pcoa_result_B_tr_T1)

# Calculate centroids for each group and timepoint
centroid_data_B_tr_T1 <- combined_data_B_tr_T1 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
combined_data_B_tr_T1$Timepoint <- factor(combined_data_B_tr_T1$Timepoint, levels = c("T0", "T1", "T2", "T3"))

# Arrange data for plotting
arranged_data_B_tr_T1 <- centroid_data_B_tr_T1 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_B_tr_T1)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_B_tr_T1)$bulk_rhizo, 
                                                   "_", sample_data(ps_B_tr_T1)$Timepoint)

# Panel A

S1_B_B_T1_3 <- ggplot(combined_data_B_tr_T1, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_B_tr_T1$values$Eigenvalues[1] / sum(pcoa_B_tr_T1$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_B_tr_T1$values$Eigenvalues[2] / sum(pcoa_B_tr_T1$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

S1_B_B_T1_3

ggsave(paste0(workdir, "plots/pcoa/S1_B_B_T1_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_B_B_T1_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/S1_B_B_T1_3_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_B_B_T1_3,
       width = 4, height = 3, dpi = 600)

### T2
### T2
ps_B_tr_T2 <- ps_filter(ps_B_tr, Timepoint == "T2")

pcoa_B_tr_T2 <- ordinate(ps_B_tr_T2, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_B_tr_T2 <- as.data.frame(pcoa_B_tr_T2$vectors)

# Extract sample data for temporal information
temporal_data_B_tr_T2 <- as.data.frame(sample_data(ps_B_tr_T2))

# Combine data for plotting
combined_data_B_tr_T2 <- cbind(temporal_data_B_tr_T2, pcoa_result_B_tr_T2)

# Calculate centroids for each group and timepoint
centroid_data_B_tr_T2 <- combined_data_B_tr_T2 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
#combined_data_B_tr_T2$Timepoint <- factor(combined_data_B_tr_T2$Timepoint, levels = c("T0", "T2", "T2", "T3"))

# Arrange data for plotting
arranged_data_B_tr_T2 <- centroid_data_B_tr_T2 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_B_tr_T2)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_B_tr_T2)$bulk_rhizo, 
                                                  "_", sample_data(ps_B_tr_T2)$Timepoint)

# Panel A
S1_B_B_T2_3 <- ggplot(combined_data_B_tr_T2, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_B_tr_T2$values$Eigenvalues[1] / sum(pcoa_B_tr_T2$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_B_tr_T2$values$Eigenvalues[2] / sum(pcoa_B_tr_T2$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

S1_B_B_T2_3

ggsave(paste0(workdir, "plots/pcoa/S1_B_B_T2_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_B_B_T2_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/S1_B_B_T2_3_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_B_B_T2_3,
       width = 4, height = 3, dpi = 600)


### T3
ps_B_tr_T3 <- ps_filter(ps_B_tr, Timepoint == "T3")

pcoa_B_tr_T3 <- ordinate(ps_B_tr_T3, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_B_tr_T3 <- as.data.frame(pcoa_B_tr_T3$vectors)

# Extract sample data for temporal information
temporal_data_B_tr_T3 <- as.data.frame(sample_data(ps_B_tr_T3))

# Combine data for plotting
combined_data_B_tr_T3 <- cbind(temporal_data_B_tr_T3, pcoa_result_B_tr_T3)

# Calculate centroids for each group and timepoint
centroid_data_B_tr_T3 <- combined_data_B_tr_T3 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
#combined_data_B_tr_T3$Timepoint <- factor(combined_data_B_tr_T3$Timepoint, levels = c("T0", "T3", "T3", "T3"))

# Arrange data for plotting
arranged_data_B_tr_T3 <- centroid_data_B_tr_T3 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_B_tr_T3)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_B_tr_T3)$bulk_rhizo, 
                                                  "_", sample_data(ps_B_tr_T3)$Timepoint)

# Panel A
S1_B_B_T3_3 <-  ggplot(combined_data_B_tr_T3, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_B_tr_T3$values$Eigenvalues[1] / sum(pcoa_B_tr_T3$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_B_tr_T3$values$Eigenvalues[2] / sum(pcoa_B_tr_T3$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()


S1_B_B_T3_3

ggsave(paste0(workdir, "plots/pcoa/S1_B_B_T3_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_B_B_T3_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/S1_B_B_T3_3_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_B_B_T3_3,
       width = 4, height = 3, dpi = 600)

```


## Panel C


```{r}

alpha_B_tr <- alpha(ps_B_tr, index = "Observed")

alpha_B_tr$SampleID <- rownames(alpha_B_tr)
alpha_B_tr_m <- merge(alpha_B_tr, meta(ps_B_tr), by = "SampleID")

alpha_B_tr_m <- alpha_B_tr_m %>%
  mutate(
    Numeric_Timepoint = case_when(
      Timepoint == "T0" ~ 0,
      Timepoint == "T1" ~ 9,
      Timepoint == "T2" ~ 14,
      Timepoint == "T3" ~ 19,
      TRUE ~ as.numeric(Timepoint)   # ensures numeric output
    )
  )


alpha_B_tr_m <- alpha_B_tr_m %>%
  mutate(
    Time_month = case_when(
      Timepoint == "T0" ~ "March",
      Timepoint == "T1" ~ "June",
      Timepoint == "T2" ~ "July",
      Timepoint == "T3" ~ "August",
      TRUE ~ Timepoint
    )
  )

alpha_B_tr_m$Time_month <- factor(alpha_B_tr_m$Time_month,
                                  levels = c("March","June","July","August"))


S1_B_C <- ggplot(alpha_B_tr_m, aes(x = as.factor(Time_month), y = observed, fill = bulk_rhizo, color = bulk_rhizo)) +
  geom_boxplot(color = "black", alpha = 0.7, position = position_dodge(0.75)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.5, dodge.width = 0.75), alpha = 1, color = "black") +
  labs(x = "", y = "Number of ASVs") +
  scale_fill_manual(name = "soil", values = compartment_colors) +
  scale_color_manual(name = "soil", values = compartment_colors) +
  theme_soil()


library(dplyr)
library(rstatix)

B_richness_stat_test <- alpha_B_tr_m %>%
  group_by(Time_month) %>%
  filter(n_distinct(bulk_rhizo) > 1) %>%
  wilcox_test(observed ~ bulk_rhizo) %>%
  add_significance("p") %>%
  add_xy_position(x = "Time_month", dodge = 0.8)

S1_B_C_stat <- S1_B_C +
  stat_pvalue_manual(
    B_richness_stat_test,             # your stat_test dataframe
    label = "p.signif",    # shows *, **, etc.
    tip.length = 0.02,     # length of the whisker
    hide.ns = FALSE,        # hide non-significant comparisons
    position = position_dodge(width = 0.8),  # align with dodged boxes
    label.size = 8
  )

S1_B_C_stat

ggsave(paste0(workdir, "plots/alpha_moisture_plot/S1_B_PanelC_stats_", Sys.Date(), ".pdf"),
       device = "pdf", plot = S1_B_C_stat,
       width = 7, height = 6, dpi = 600)

```


## Fungi

## Panel A - Bacteria

```{r}
###########################################################################################
# transform the data
ps_F_tr <- microbiome::transform(physeq_F, 'hellinger')

# PCoA - with Bray Curtis
pcoa_F_tr <- ordinate(ps_F_tr, "PCoA", distance = "bray", binary = FALSE)

# Extract PCoA results
pcoa_result_F_tr <- as.data.frame(pcoa_F_tr$vectors)

# Extract sample data for temporal information
temporal_data_F_tr <- as.data.frame(sample_data(ps_F_tr))

# Combine
combined_data_F_tr <- cbind(temporal_data_F_tr, pcoa_result_F_tr)

# Centroids
centroid_data_F_tr <- combined_data_F_tr %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Factor order for time
combined_data_F_tr$Timepoint <- factor(combined_data_F_tr$Timepoint,
                                       levels = c("T0","T1","T2","T3"))

arranged_data_F_tr <- centroid_data_F_tr %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_F_tr)$bulk_rhizo_t_2 <- paste0(sample_data(ps_F_tr)$bulk_rhizo, "_",
                                              sample_data(ps_F_tr)$Timepoint)

# Panel A
S1_F_A <- 
 ggplot(combined_data_F_tr, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo_t_2), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_F_tr$values$Eigenvalues[1] / sum(pcoa_F_tr$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_F_tr$values$Eigenvalues[2] / sum(pcoa_F_tr$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()  


S1_F_A

ggsave(paste0(workdir, "plots/pcoa/S1_F_PanelA_", Sys.Date(), ".pdf"),
       device = "pdf", plot = S1_F_A, width = 7, height = 6, dpi = 600)

```


## Panel B

```{r}

# PCoA - with Bray Curtis

### T1
ps_F_tr_T1 <- ps_filter(ps_F_tr, Timepoint == "T1")

pcoa_F_tr_T1 <- ordinate(ps_F_tr_T1, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_F_tr_T1 <- as.data.frame(pcoa_F_tr_T1$vectors)

# Extract sample data for temporal information
temporal_data_F_tr_T1 <- as.data.frame(sample_data(ps_F_tr_T1))

# Combine data for plotting
combined_data_F_tr_T1 <- cbind(temporal_data_F_tr_T1, pcoa_result_F_tr_T1)

# Calculate centroids for each group and timepoint
centroid_data_F_tr_T1 <- combined_data_F_tr_T1 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
combined_data_F_tr_T1$Timepoint <- factor(combined_data_F_tr_T1$Timepoint, levels = c("T0", "T1", "T2", "T3"))

# Arrange data for plotting
arranged_data_F_tr_T1 <- centroid_data_F_tr_T1 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_F_tr_T1)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_F_tr_T1)$bulk_rhizo, 
                                                  "_", sample_data(ps_F_tr_T1)$Timepoint)

# Panel A
S1_F_B_T1_3 <- ggplot(combined_data_F_tr_T1, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_F_tr_T1$values$Eigenvalues[1] / sum(pcoa_F_tr_T1$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_F_tr_T1$values$Eigenvalues[2] / sum(pcoa_F_tr_T1$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

S1_F_B_T1_3

ggsave(paste0(workdir, "plots/pcoa/S1_F_B_T1_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_F_B_T1_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/S1_F_B_T1_3_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_F_B_T1_3,
       width = 4, height = 3, dpi = 600)

### T2
### T2
ps_F_tr_T2 <- ps_filter(ps_F_tr, Timepoint == "T2")

pcoa_F_tr_T2 <- ordinate(ps_F_tr_T2, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_F_tr_T2 <- as.data.frame(pcoa_F_tr_T2$vectors)

# Extract sample data for temporal information
temporal_data_F_tr_T2 <- as.data.frame(sample_data(ps_F_tr_T2))

# Combine data for plotting
combined_data_F_tr_T2 <- cbind(temporal_data_F_tr_T2, pcoa_result_F_tr_T2)

# Calculate centroids for each group and timepoint
centroid_data_F_tr_T2 <- combined_data_F_tr_T2 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
#combined_data_F_tr_T2$Timepoint <- factor(combined_data_F_tr_T2$Timepoint, levels = c("T0", "T2", "T2", "T3"))

# Arrange data for plotting
arranged_data_F_tr_T2 <- centroid_data_F_tr_T2 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_F_tr_T2)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_F_tr_T2)$bulk_rhizo, 
                                                  "_", sample_data(ps_F_tr_T2)$Timepoint)

# Panel A
S1_F_B_T2_3 <- ggplot(combined_data_F_tr_T2, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_F_tr_T2$values$Eigenvalues[1] / sum(pcoa_F_tr_T2$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_F_tr_T2$values$Eigenvalues[2] / sum(pcoa_F_tr_T2$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

S1_F_B_T2_3

ggsave(paste0(workdir, "plots/pcoa/S1_F_B_T2_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_F_B_T2_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/S1_F_B_T2_3_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_F_B_T2_3,
       width = 4, height = 3, dpi = 600)


### T3
ps_F_tr_T3 <- ps_filter(ps_F_tr, Timepoint == "T3")

pcoa_F_tr_T3 <- ordinate(ps_F_tr_T3, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_F_tr_T3 <- as.data.frame(pcoa_F_tr_T3$vectors)

# Extract sample data for temporal information
temporal_data_F_tr_T3 <- as.data.frame(sample_data(ps_F_tr_T3))

# Combine data for plotting
combined_data_F_tr_T3 <- cbind(temporal_data_F_tr_T3, pcoa_result_F_tr_T3)

# Calculate centroids for each group and timepoint
centroid_data_F_tr_T3 <- combined_data_F_tr_T3 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
#combined_data_F_tr_T3$Timepoint <- factor(combined_data_F_tr_T3$Timepoint, levels = c("T0", "T3", "T3", "T3"))

# Arrange data for plotting
arranged_data_F_tr_T3 <- centroid_data_F_tr_T3 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_F_tr_T3)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_F_tr_T3)$bulk_rhizo, 
                                                  "_", sample_data(ps_F_tr_T3)$Timepoint)

# Panel A
S1_F_B_T3_3 <- ggplot(combined_data_F_tr_T3, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_F_tr_T3$values$Eigenvalues[1] / sum(pcoa_F_tr_T3$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_F_tr_T3$values$Eigenvalues[2] / sum(pcoa_F_tr_T3$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

S1_F_B_T3_3

ggsave(paste0(workdir, "plots/pcoa/S1_F_B_T3_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_F_B_T3_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/S1_F_B_T3_3_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S1_F_B_T3_3,
       width = 4, height = 3, dpi = 600)


```


## Panel C


```{r}

alpha_F_tr <- alpha(ps_F_tr, index = "Observed")

alpha_F_tr$SampleID <- rownames(alpha_F_tr)
alpha_F_tr_m <- merge(alpha_F_tr, meta(ps_F_tr), by = "SampleID")

alpha_F_tr_m <- alpha_F_tr_m %>%
  mutate(
    Numeric_Timepoint = case_when(
      Timepoint == "T0" ~ 0,
      Timepoint == "T1" ~ 9,
      Timepoint == "T2" ~ 14,
      Timepoint == "T3" ~ 19,
      TRUE ~ as.numeric(Timepoint)   # ensures numeric output
    )
  )


alpha_F_tr_m <- alpha_F_tr_m %>%
  mutate(
    Time_month = case_when(
      Timepoint == "T0" ~ "March",
      Timepoint == "T1" ~ "June",
      Timepoint == "T2" ~ "July",
      Timepoint == "T3" ~ "August",
      TRUE ~ Timepoint
    )
  )

alpha_F_tr_m$Time_month <- factor(alpha_F_tr_m$Time_month,
                                  levels = c("March","June","July","August"))


S1_F_C <- ggplot(alpha_F_tr_m, aes(x = as.factor(Time_month), y = observed, fill = bulk_rhizo, color = bulk_rhizo)) +
  geom_boxplot(color = "black", alpha = 0.7, position = position_dodge(0.75)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.5, dodge.width = 0.75), alpha = 1, color = "black") +
  labs(x = "", y = "Number of ASVs") +
  scale_fill_manual(name = "soil", values = compartment_colors) +
  scale_color_manual(name = "soil", values = compartment_colors) +
  theme_soil()


library(dplyr)
library(rstatix)

F_richness_stat_test <- alpha_F_tr_m %>%
  group_by(Time_month) %>%
  filter(n_distinct(bulk_rhizo) > 1) %>%
  wilcox_test(observed ~ bulk_rhizo) %>%
  add_significance("p") %>%
  add_xy_position(x = "Time_month", dodge = 0.8)

S1_F_C_stat <- S1_F_C +
  stat_pvalue_manual(
    F_richness_stat_test,             # your stat_test dataframe
    label = "p.signif",    # shows *, **, etc.
    tip.length = 0.02,     # length of the whisker
    hide.ns = FALSE,        # hide non-significant comparisons
    position = position_dodge(width = 0.8),  # align with dodged boxes
    label.size = 8
  )

S1_F_C_stat


ggsave(paste0(workdir, "plots/alpha_moisture_plot/S1_F_PanelC_stats_", Sys.Date(), ".pdf"),
       device = "pdf", plot = S1_F_C_stat,
       width = 7, height = 6, dpi = 600)

```


## Combined Figure S1

```{r}

S1_panel_CD <- (S1_B_B_T1_3 + ggtitle("Bacteria - Timepoint 1 - June") | 
                S1_F_B_T1_3 + ggtitle("Fungi - Timepoint 1 - June")) /
               (S1_B_B_T2_3 + ggtitle("Bacteria - Timepoint 2 - July") | 
                S1_F_B_T2_3 + ggtitle("Fungi - Timepoint 2 - July")) /
               (S1_B_B_T3_3 + ggtitle("Bacteria - Timepoint 3 - August") | 
                S1_F_B_T3_3 + ggtitle("Fungi - Timepoint 3 - August")) +
  plot_layout(guides = "collect") & 
  theme(plot.margin = margin(1,1,1,1))


S1_panel_CD

# Combine with patchwork
left_column <- (S1_B_A + ggtitle("Bacterial community composition")) / (S1_F_A + ggtitle("Fungal community composition"))
right_column <- S1_B_C  / S1_F_C

Fig_S1_combined <- left_column | S1_panel_CD | right_column +
  plot_annotation(tag_levels = 'A')

Fig_S1_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Fig_S1_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig_S1_combined, 
       width = 20, height = 14, dpi = 600)


```


# Figure S2


```{r}

packageVersion("eulerr")

today <- format(Sys.Date(), "%Y-%m-%d")

# Dataset prefixes
prefixes_euler <- c("DV", "B", "F")

# Colors
colors_euler <- c("bulk" = "#BA7F3A", "rhizo" = "#3C8E3A")

# Loop over each prefix and generate plot
for (prefix in prefixes_euler) {
  # Construct object names for vectors
  bulk_var <- get(paste0("venn_", prefix, "_tr_bulk"))
  rhizo_var <- get(paste0("venn_", prefix, "_tr_rhizo"))
  
  # Compute intersections
  only_bulk <- length(setdiff(bulk_var, rhizo_var))
  only_rhizo <- length(setdiff(rhizo_var, bulk_var))
  shared <- length(intersect(bulk_var, rhizo_var))
  
  # Create Euler diagram data
  venn_data <- euler(c(
    "Bulk" = only_bulk,
    "Rhizo" = only_rhizo,
    "Bulk&Rhizo" = shared
  ))
  
  # Plot and save to file
  p <- plot(
    venn_data,
    fills = list(fill = c(colors_euler["bulk"], colors_euler["rhizo"]), alpha = 0.7),
    edges = list(lwd = 2),
    labels = list(font = 2),
    quantities = TRUE,
    main = NULL  # No title
  )
  
  # Save plot
  ggsave(
    filename = paste0(paste0(workdir, "plots/euler/"), "euler_", prefix, "_", Sys.Date(), ".pdf"),
    plot = p,
    width = 5, height = 5
  )
}

```


# Figure S3

Bulk soil plot location and AMF

## Panel A

### DV viruses
Bulk soil plot location

```{r}

S3_DV_A <- ggplot(combined_data_DV_tr_bulk, aes(x = Axis.1, y = Axis.2, color = Plot, shape = Plot)) +
  geom_point(aes(fill = Plot), size = 6, color = "black") +
  geom_convexhull(aes(color = Plot,fill = Plot, group = Plot), alpha = 0.1) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_bulk$values$Eigenvalues[1] / sum(pcoa_DV_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_bulk$values$Eigenvalues[2] / sum(pcoa_DV_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() + labs(title = "Bulk soil DNA viral community")

S3_DV_A

ggsave(paste0(workdir, "plots/pcoa/pcoa_DV_bulk_plots_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S3_DV_A,
       width = 7, height = 6, dpi = 600)


```



### Bacteria
Bulk soil plot location 

```{r}

S3_B_A <- ggplot(combined_data_B_tr_bulk, aes(x = Axis.1, y = Axis.2, color = Plot, shape = Plot)) +
  geom_point(aes(fill = Plot), size = 6, color = "black") +
  geom_convexhull(aes(color = Plot,fill = Plot, group = Plot), alpha = 0.1) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_B_tr_bulk$values$Eigenvalues[1] / sum(pcoa_B_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_B_tr_bulk$values$Eigenvalues[2] / sum(pcoa_B_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() + labs(title = "Bulk soil bacterial community")

S3_B_A

ggsave(paste0(workdir, "plots/pcoa/pcoa_B_bulk_plots_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S3_B_A,
       width = 7, height = 6, dpi = 600)

```

### Fungi

```{r}

S3_F_A <- ggplot(combined_data_F_tr_bulk, aes(x = Axis.1, y = Axis.2, color = Plot, shape = Plot)) +
  geom_point(aes(fill = Plot), size = 6, color = "black") +
  geom_convexhull(aes(color = Plot,fill = Plot, group = Plot), alpha = 0.1) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_F_tr_bulk$values$Eigenvalues[1] / sum(pcoa_F_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_F_tr_bulk$values$Eigenvalues[2] / sum(pcoa_F_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() + labs(title = "Bulk soil fungal community")

S3_F_A

ggsave(paste0(workdir, "plots/pcoa/pcoa_F_bulk_plots_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S3_F_A,
       width = 7, height = 6, dpi = 600)
```

## Panel B

AMF treatment

### DV

```{r}

S3_DV_B <- ggplot(combined_data_DV_tr_bulk, aes(x = Axis.1, y = Axis.2, color = Treat, shape = Treat)) +
  geom_point(aes(fill = Treat), size = 6, color = "black") +
  geom_convexhull(aes(color = Treat,fill = Treat, group = Treat), alpha = 0.1) +
  scale_color_manual(values = amf_colors) +
  scale_fill_manual(values = amf_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_bulk$values$Eigenvalues[1] / sum(pcoa_DV_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_bulk$values$Eigenvalues[2] / sum(pcoa_DV_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() + labs(title = "Bulk soil viral community")

S3_DV_B

ggsave(paste0(workdir, "plots/pcoa/pcoa_DV_bulk_AMFtreatment_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S3_DV_B,
       width = 7, height = 6, dpi = 600)

```


### Bacteria

```{r}

S3_B_B <- ggplot(combined_data_B_tr_bulk, aes(x = Axis.1, y = Axis.2, color = Treat, shape = Treat)) +
  geom_point(aes(fill = Treat), size = 6, color = "black") +
  geom_convexhull(aes(color = Treat,fill = Treat, group = Treat), alpha = 0.1) +
  scale_color_manual(values = amf_colors) +
  scale_fill_manual(values = amf_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_B_tr_bulk$values$Eigenvalues[1] / sum(pcoa_B_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_B_tr_bulk$values$Eigenvalues[2] / sum(pcoa_B_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() + labs(title = "Bulk soil bacterial community")

S3_B_B

ggsave(paste0(workdir, "plots/pcoa/pcoa_B_bulk_AMFtreatment_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S3_B_B,
       width = 7, height = 6, dpi = 600)

```


### Fungi

```{r}

S3_F_B <- ggplot(combined_data_F_tr_bulk, aes(x = Axis.1, y = Axis.2, color = Treat, shape = Treat)) +
  geom_point(aes(fill = Treat), size = 6, color = "black") +
  geom_convexhull(aes(color = Treat,fill = Treat, group = Treat), alpha = 0.1) +
  scale_color_manual(values = amf_colors) +
  scale_fill_manual(values = amf_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_F_tr_bulk$values$Eigenvalues[1] / sum(pcoa_F_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_F_tr_bulk$values$Eigenvalues[2] / sum(pcoa_F_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() + labs(title = "Bulk soil fungal community")

S3_F_B

ggsave(paste0(workdir, "plots/pcoa/pcoa_F_bulk_AMFtreatment_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = S3_F_B,
       width = 7, height = 6, dpi = 600)

```

## Combined Fig S3


```{r}
# Combine with patchwork
Figure_S3_combined <- (S3_DV_A | S3_DV_B  ) / (S3_B_A | S3_B_B  ) / (S3_F_A | S3_F_B  ) +
  plot_annotation(tag_levels = 'A')

Figure_S3_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_S3_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_S3_combined, 
       width = 14, height = 18, dpi = 600)
```


# Figure S4

Bacteria and Fungi - Bulk soil moisture

## Panel A 

Bacteria
```{r}
## ---------------------------
## Split to bulk compartment
## ---------------------------

ps_B_tr_bulk <- subset_samples(ps_B_tr, bulk_rhizo == "bulk")
ps_B_tr_bulk <- filter_taxa(ps_B_tr_bulk, function(x) sum(x) > 0, TRUE)

## Timepoint numeric
ps_B_tr_bulk@sam_data$Numeric_Timepoint <- case_when(
  ps_B_tr_bulk@sam_data$Timepoint == "T0" ~ 0,
  ps_B_tr_bulk@sam_data$Timepoint == "T1" ~ 9,
  ps_B_tr_bulk@sam_data$Timepoint == "T2" ~ 14,
  ps_B_tr_bulk@sam_data$Timepoint == "T3" ~ 19
)


ps_B_tr_bulk@sam_data$Time_weeks <- case_when(
  ps_B_tr_bulk@sam_data$Timepoint == "T0" ~ 0,
  ps_B_tr_bulk@sam_data$Timepoint == "T1" ~ 9,
  ps_B_tr_bulk@sam_data$Timepoint == "T2" ~ 14,
  ps_B_tr_bulk@sam_data$Timepoint == "T3" ~ 19
)


ps_B_tr_bulk@sam_data$Time_month <- case_when(
  ps_B_tr_bulk@sam_data$Timepoint == "T0" ~ "March",
  ps_B_tr_bulk@sam_data$Timepoint == "T1" ~ "June",
  ps_B_tr_bulk@sam_data$Timepoint == "T2" ~ "July",
  ps_B_tr_bulk@sam_data$Timepoint == "T3" ~ "August"
)

pcoa_B_tr_bulk <- ordinate(ps_B_tr_bulk, "PCoA", distance = "bray", binary = FALSE)

## Prepare data
pcoa_result_B_tr_bulk <- as.data.frame(pcoa_B_tr_bulk$vectors)
temporal_data_B_tr_bulk <- as.data.frame(sample_data(ps_B_tr_bulk))
combined_data_B_tr_bulk <- cbind(temporal_data_B_tr_bulk, pcoa_result_B_tr_bulk)

centroid_data_B_tr_bulk <- combined_data_B_tr_bulk %>%
  group_by(Timepoint) %>%
  summarize(
    Centroid_X = mean(Axis.1),
    Centroid_Y = mean(Axis.2)
  )

combined_data_B_tr_bulk$Timepoint <- factor(
  combined_data_B_tr_bulk$Timepoint,
  levels = c("T0", "T1", "T2", "T3")
)

## Supplemental Figure 3A – Bacteria
S4_B_A <- ggplot(combined_data_B_tr_bulk, aes(x = Axis.1, y = Axis.2)) +
  geom_point(aes(fill = soil_moisture_perc, shape = Time_month),
             color = "black", size = 6) +
  scale_fill_gradient(low = "#D0B17F", high = "#2171b5") +
  scale_shape_manual(
    breaks = c("March", "June", "July", "August"),
    values = c(21, 22, 23, 24)
  ) +
  geom_text_repel(
    data = centroid_data_B_tr_bulk,
    aes(x = Centroid_X, y = Centroid_Y, label = Timepoint),
    size = 4, fontface = 3
  ) +
  geom_segment(
    data = centroid_data_B_tr_bulk,
    aes(
      x = lag(Centroid_X), y = lag(Centroid_Y),
      xend = Centroid_X, yend = Centroid_Y
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    color = "grey",
    na.rm = TRUE
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 16),
    panel.grid = element_blank(),
    legend.position = "none"
  ) +
  xlab(paste0("PCoA axis 1 (",
              round(pcoa_B_tr_bulk$values$Eigenvalues[1] /
                    sum(pcoa_B_tr_bulk$values$Eigenvalues) * 100, 2),
              "%)")) +
  ylab(paste0("PCoA axis 2 (",
              round(pcoa_B_tr_bulk$values$Eigenvalues[2] /
                    sum(pcoa_B_tr_bulk$values$Eigenvalues) * 100, 2),
              "%)")) +
  theme_soil()

ggsave(
  paste0(workdir, "plots/pcoa/S4_B_A_moisture_pcoa_bulk_", Sys.Date(), ".pdf"),
  plot = S4_B_A,
  width = 7, height = 6, dpi = 600
)

```


Fungi

```{r}
## ---------------------------
## Split to bulk compartment
## ---------------------------

ps_F_tr_bulk <- subset_samples(ps_F_tr, bulk_rhizo == "bulk")
ps_F_tr_bulk <- filter_taxa(ps_F_tr_bulk, function(x) sum(x) > 0, TRUE)

## Timepoint numeric
ps_F_tr_bulk@sam_data$Numeric_Timepoint <- case_when(
  ps_F_tr_bulk@sam_data$Timepoint == "T0" ~ 0,
  ps_F_tr_bulk@sam_data$Timepoint == "T1" ~ 9,
  ps_F_tr_bulk@sam_data$Timepoint == "T2" ~ 14,
  ps_F_tr_bulk@sam_data$Timepoint == "T3" ~ 19
)

ps_F_tr_bulk@sam_data$Time_weeks <- case_when(
  ps_F_tr_bulk@sam_data$Timepoint == "T0" ~ 0,
  ps_F_tr_bulk@sam_data$Timepoint == "T1" ~ 9,
  ps_F_tr_bulk@sam_data$Timepoint == "T2" ~ 14,
  ps_F_tr_bulk@sam_data$Timepoint == "T3" ~ 19
)

ps_F_tr_bulk@sam_data$Time_month <- case_when(
  ps_F_tr_bulk@sam_data$Timepoint == "T0" ~ "March",
  ps_F_tr_bulk@sam_data$Timepoint == "T1" ~ "June",
  ps_F_tr_bulk@sam_data$Timepoint == "T2" ~ "July",
  ps_F_tr_bulk@sam_data$Timepoint == "T3" ~ "August"
)

pcoa_F_tr_bulk <- ordinate(ps_F_tr_bulk, "PCoA", distance = "bray", binary = FALSE)

## Prepare data
pcoa_result_F_tr_bulk <- as.data.frame(pcoa_F_tr_bulk$vectors)
temporal_data_F_tr_bulk <- as.data.frame(sample_data(ps_F_tr_bulk))
combined_data_F_tr_bulk <- cbind(temporal_data_F_tr_bulk, pcoa_result_F_tr_bulk)

centroid_data_F_tr_bulk <- combined_data_F_tr_bulk %>%
  group_by(Timepoint) %>%
  summarize(
    Centroid_X = mean(Axis.1),
    Centroid_Y = mean(Axis.2)
  )

combined_data_F_tr_bulk$Timepoint <- factor(
  combined_data_F_tr_bulk$Timepoint,
  levels = c("T0", "T1", "T2", "T3")
)

## Supplemental Figure 3A – Fungi
S4_F_A <- ggplot(combined_data_F_tr_bulk, aes(x = Axis.1, y = Axis.2)) +
  geom_point(aes(fill = soil_moisture_perc, shape = Time_month),
             color = "black", size = 6) +
  scale_fill_gradient(low = "#D0B17F", high = "#2171b5") +
  scale_shape_manual(
    breaks = c("March", "June", "July", "August"),
    values = c(21, 22, 23, 24)
  ) +
  geom_text_repel(
    data = centroid_data_F_tr_bulk,
    aes(x = Centroid_X, y = Centroid_Y, label = Timepoint),
    size = 4, fontface = 3
  ) +
  geom_segment(
    data = centroid_data_F_tr_bulk,
    aes(
      x = lag(Centroid_X), y = lag(Centroid_Y),
      xend = Centroid_X, yend = Centroid_Y
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    color = "grey",
    na.rm = TRUE
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 16),
    panel.grid = element_blank(),
    legend.position = "none"
  ) +
  xlab(paste0("PCoA axis 1 (",
              round(pcoa_F_tr_bulk$values$Eigenvalues[1] /
                    sum(pcoa_F_tr_bulk$values$Eigenvalues) * 100, 2),
              "%)")) +
  ylab(paste0("PCoA axis 2 (",
              round(pcoa_F_tr_bulk$values$Eigenvalues[2] /
                    sum(pcoa_F_tr_bulk$values$Eigenvalues) * 100, 2),
              "%)")) +
  theme_soil()

ggsave(
  paste0(workdir, "plots/pcoa/S4_F_A_moisture_pcoa_bulk_", Sys.Date(), ".pdf"),
  plot = S4_F_A,
  width = 7, height = 6, dpi = 600
)

```


## Panel B

### Bacteria
```{r}

## --------------------------------------------
## Axis.1 vs Time — Bacteria (Supplemental)
## --------------------------------------------

data_B_bulk <- combined_data_B_tr_bulk %>%
  filter(!is.na(Time_weeks) & !is.na(Axis.1))

cor_B_time <- cor.test(data_B_bulk$Time_weeks, data_B_bulk$Axis.1)

S4_B_B <- 
  ggplot(data_B_bulk, aes(y = Axis.1, x = Time_weeks, fill = soil_moisture_perc, shape = Time_month)) +
    geom_point(size = 6) +
      geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  scale_fill_gradient(low = "#D0B17F", high = "#2171b5") +
  scale_shape_manual(
    breaks = c("March", "June", "July", "August"),
    values = c(21, 22, 23, 24)
  ) +
  annotate("text",
           y = min(data_B_bulk$Time_weeks, na.rm = TRUE),
           x = max(data_B_bulk$Axis.1, na.rm = TRUE),
           label = paste0("r = ", round(cor_B_time$estimate, 2),
                          ", p = ", signif(cor_B_time$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 6) +
  labs(y = "Bulk soil bacterial PCoA Axis 1", x = "Time (weeks)", color = "Soil moisture (%)") + 
  theme_soil()

S4_B_B

ggsave(
  paste0(workdir, "plots/axis_moisture/S4_B_B_Axis1_vs_Time_", Sys.Date(), ".pdf"),
  plot = S4_B_B,
  width = 7, height = 6, dpi = 600
)


```


### Fungi
```{r}
## --------------------------------------------
##  Axis.1 vs Time — Fungi (Supplemental)
## --------------------------------------------

data_F_bulk <- combined_data_F_tr_bulk %>%
  filter(!is.na(Time_weeks) & !is.na(Axis.1))

cor_F_time <- cor.test(data_F_bulk$Time_weeks, data_F_bulk$Axis.1)

S4_F_B <- ggplot(data_F_bulk, aes(x = Time_weeks, y = Axis.1, color = soil_moisture_perc)) +
  geom_point(size = 6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  annotate("text",
           x = min(data_F_bulk$Time_weeks, na.rm = TRUE),
           y = max(data_F_bulk$Axis.1, na.rm = TRUE),
           label = paste0("r = ", round(cor_F_time$estimate, 2),
                          ", p = ", signif(cor_F_time$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 4) +
  scale_color_gradient(
    low = "#C9A46C",
    high = "#2171b5"
  ) +
  labs(x = "Time (weeks)", y = "Axis 1", color = "Soil moisture (%)") + 
  theme_soil()

S4_F_B

ggsave(
  paste0(workdir, "plots/axis_moisture/S4_F_B_Axis1_vs_Time_", Sys.Date(), ".pdf"),
  plot = S4_F_B,
  width = 7, height = 6, dpi = 600
)


```


## Panel C

Axis 1 and soil moisture

### Bacteria

```{r}

## Axis.1 vs Soil Moisture (Panel C)

data_b_bulk <- combined_data_B_tr_bulk %>%
  filter(!is.na(soil_moisture_perc) & !is.na(Axis.1))

cor_b_soil <- cor.test(data_b_bulk$soil_moisture_perc, data_b_bulk$Axis.1)

S4_B_C <- ggplot(data_B_bulk, aes(y = Axis.1, x = soil_moisture_perc, fill = soil_moisture_perc, shape = Time_month)) +
       geom_point(size = 6) +
         geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
         scale_fill_gradient(low = "#D0B17F", high = "#2171b5") +
         scale_shape_manual(
           breaks = c("March", "June", "July", "August"),
           values = c(21, 22, 23, 24)
         ) +
  annotate("text",
           x = min(data_b_bulk$soil_moisture_perc, na.rm = TRUE),
           y = max(data_b_bulk$Axis.1, na.rm = TRUE),
           label = paste0("r = ", round(cor_b_soil$estimate, 2),
                          ", p = ", signif(cor_b_soil$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 6) +
  labs(y = "Bulk soil bacterial PCoA Axis 1", x = "Soil moisture (%)", color = "Soil moisture (%)") + 
  theme_soil()


S4_B_C

ggsave(
  paste0(workdir, "plots/axis_moisture/B_Axis1_vs_SoilMoisture_", Sys.Date(), ".pdf"),
  plot = S4_B_C,
  width = 7, height = 6,
  dpi = 600
)

```


### Fungi

```{r}

## Axis.1 vs Soil Moisture (Panel C)

data_f_bulk <- combined_data_F_tr_bulk %>%
  filter(!is.na(soil_moisture_perc) & !is.na(Axis.1))

cor_f_soil <- cor.test(data_f_bulk$soil_moisture_perc, data_f_bulk$Axis.1)

S4_F_C <- ggplot(data_f_bulk, 
                 aes(x = soil_moisture_perc, 
                     y = Axis.1, 
                     color = Time_month)) +
  geom_point(size = 5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  annotate("text",
           x = min(data_f_bulk$soil_moisture_perc, na.rm = TRUE),
           y = max(data_f_bulk$Axis.1, na.rm = TRUE),
           label = paste0("r = ", round(cor_f_soil$estimate, 2),
                          ", p = ", signif(cor_f_soil$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 4) +
  scale_color_manual(values = c(
    "March"  = "#C9A46C",
    "June"   = "#BA7F3A",
    "July"   = "#8F5A28",
    "August" = "#5E4224"
  )) +
  labs(x = "Soil moisture (%)",
       y = "Axis 1",
       color = "Time (month)") +
  theme_soil()

S4_F_C

ggsave(
  paste0(workdir, "plots/axis_moisture/F_Axis1_vs_SoilMoisture_", Sys.Date(), ".pdf"),
  plot = S4_F_C,
  width = 7, height = 6,
  dpi = 600
)

```

#### Fungi - Axis 2

```{r}


cor_f_soil_A2 <- cor.test(data_f_bulk$soil_moisture_perc, data_f_bulk$Axis.2)

S4_F_C2 <- ggplot(data_f_bulk, aes(y = Axis.2, x = soil_moisture_perc, fill = soil_moisture_perc, shape = Time_month)) +
       geom_point(size = 6) +
         geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
         scale_fill_gradient(low = "#D0B17F", high = "#2171b5") +
         scale_shape_manual(
           breaks = c("March", "June", "July", "August"),
           values = c(21, 22, 23, 24)
         ) +
  annotate("text",
           x = min(data_f_bulk$soil_moisture_perc, na.rm = TRUE),
           y = max(data_f_bulk$Axis.2, na.rm = TRUE),
           label = paste0("r = ", round(cor_f_soil_A2$estimate, 2),
                          ", p = ", signif(cor_f_soil_A2$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 6) +
  labs(y = "Bulk soil fungal PCoA Axis 2", x = "Soil moisture (%)", color = "Soil moisture (%)") + 
  theme_soil()

S4_F_C2

ggsave(
  paste0(workdir, "plots/axis_moisture/F_Axis2_vs_SoilMoisture_", Sys.Date(), ".pdf"),
  plot = S4_F_C2,
  width = 7, height = 6,
  dpi = 600
)

```


## Panel D


### Bacteria

```{r}

## Bacteria
S4_B_D <- 
  
ggplot(alpha_B_tr_m %>% filter( bulk_rhizo == "bulk"),
       aes(x = soil_moisture_perc, y = observed, fill = soil_moisture_perc, shape = Time_month)
) +
  geom_point(size = 6) +
  labs(
    title = "",
    x = "Soil moisture (%)",
    y = "Number of bacterial ASVs"
  ) +
  
  # Single regression line for all bulk samples
  geom_smooth(
    data = alpha_B_tr_m %>% filter( bulk_rhizo == "bulk"),
    mapping = aes(x = soil_moisture_perc, y = observed), # explicitly set
    method = "lm",
    se = FALSE,
    color = "black",
    inherit.aes = FALSE # prevents re-using main plot aesthetics
  ) +
  
  # R² and p-value for all bulk samples
  stat_poly_eq(
    data = alpha_B_tr_m %>% filter( bulk_rhizo == "bulk"),
    mapping = aes(x = soil_moisture_perc, y = observed,
                  label = paste(..rr.label.., ..p.value.label.., sep = "~~~")),
    formula = y ~ x,
    parse = TRUE,
    color = "black",
    inherit.aes = FALSE
  ) +
  
  # Manual color and shape scales
  moisture_gradient_fill +
  scale_shape_manual(
    name = "Time_month",
    breaks = c("March", "June", "July", "August"),
    values = c(21,22,23,24)
  ) +
  theme_soil()


S4_B_D

ggsave(
  filename = paste0(workdir, "plots/richness_moisture/scatter_alpha_moisture_plot_B_time_", Sys.Date(), ".pdf"),
  plot = S4_B_D,
  width = 7, height = 6
)



```

### Fungi

```{r}
## Fungi
S4_F_D <- 
  
ggplot(alpha_F_tr_m %>% filter( bulk_rhizo == "bulk"),
       aes(x = soil_moisture_perc, y = observed, fill = soil_moisture_perc, shape = Time_month)
) +
  geom_point(size = 6) +
  labs(
    title = "",
    x = "Soil moisture (%)",
    y = "Number of fungal ASVs"
  ) +
  
  # Single regression line for all bulk samples
  geom_smooth(
    data = alpha_F_tr_m %>% filter( bulk_rhizo == "bulk"),
    mapping = aes(x = soil_moisture_perc, y = observed), # explicitly set
    method = "lm",
    se = FALSE,
    color = "black",
    inherit.aes = FALSE # prevents re-using main plot aesthetics
  ) +
  
  # R² and p-value for all bulk samples
  stat_poly_eq(
    data = alpha_F_tr_m %>% filter( bulk_rhizo == "bulk"),
    mapping = aes(x = soil_moisture_perc, y = observed,
                  label = paste(..rr.label.., ..p.value.label.., sep = "~~~")),
    formula = y ~ x,
    parse = TRUE,
    color = "black",
    inherit.aes = FALSE
  ) +
  
  # Manual color and shape scales
  moisture_gradient_fill +
  scale_shape_manual(
    name = "Time_month",
    breaks = c("March", "June", "July", "August"),
    values = c(21,22,23,24)
  ) +
  theme_soil()

S4_F_D

ggsave(
  filename = paste0(workdir, "plots/richness_moisture/scatter_alpha_moisture_plot_F_time_", Sys.Date(), ".pdf"),
  plot = S4_F_D,
  width = 7, height = 6
)

```


## Figure S4 - G

PCA of bulk soil chemistry

```{r}

ordered_sample_ids <- sample_data(ps_DV_tr_bulk)$SampleID

virus_data_ordered <- virus_data[ordered_sample_ids, ]

# Scale and Order Soil Chemistry Data
soil_chem_data <- meta(ps_DV_tr_bulk)
soil_chem_data <- soil_chem_data[ordered_sample_ids, ]

## WITH soil moisture
soil_chem_data_scaled <- soil_chem_data %>% 
  as.data.frame() %>% 
  select(-('WDRF_Buffer_pH')) %>%
  mutate(across(`soil_moisture_perc`:`CEC_Sum_of_Cations_me_100g`, ~scale(.))) %>%
  select(`soil_moisture_perc`:`CEC_Sum_of_Cations_me_100g`) 

soil_chem_data_scaled$SampleID <- rownames(soil_chem_data_scaled)

soil_chem_data_scaled_NO_moisture <- soil_chem_data_scaled %>%
  select(-soil_moisture_perc)   # 

# Run PCA
pca_chem <- prcomp(soil_chem_data_scaled_NO_moisture, center = TRUE, scale. = TRUE)

# Summary of PCA
summary(pca_chem)  # variance explained by PC1, PC2, etc.

# Extract scores for samples
scores <- as.data.frame(pca_chem$x)

# Add metadata (time and soil moisture)
# If there is a column for bulk/rhizo, filter bulk only
temporal_df_bulk <- temporal_data_DV_tr[temporal_data_DV_tr$bulk_rhizo == "bulk", ]

# Now combine with PCA scores
pca_soil_chem_data <- cbind(temporal_df_bulk, scores)

######################
# Make Time_month a factor
pca_soil_chem_data$Time_month <- factor(pca_soil_chem_data$Time_month, levels = c("March", "June", "July", "August"))

# Basic PCA biplot with convex hulls by Time_month
S4_G <- 
ggplot(pca_soil_chem_data, aes(x = PC1, y = PC2, fill = soil_moisture_perc, shape = Time_month)) +
  geom_point(size = 6) +
  scale_fill_gradient(low = "#D0B17F", high = "#2171b5") +
  scale_shape_manual(
    breaks = c("March", "June", "July", "August"),
    values = c(21, 22, 23, 24)
  ) +
  xlab(paste0("PC1 (", round(summary(pca_chem)$importance[2,1]*100, 1), "%)")) +
  ylab(paste0("PC2 (", round(summary(pca_chem)$importance[2,2]*100, 1), "%)")) +
    labs(y = "Bulk soil chemistry PCA Axis 2", 
         x = "Bulk soil chemistry PCA Axis 1", 
         fill = "Soil moisture (%)") +
    theme_soil()


S4_G

ggsave(paste0(workdir, "plots/axis_moisture/PCA_BulkChemistry_", Sys.Date(), ".pdf"),
       plot = S4_G, 
       width = 7, height = 6, 
       dpi = 600)


```

## Combine Fig S4 ABCD

```{r}
# Combine with patchwork
library(patchwork)

Fig_S4_combined <- (S4_B_C | S4_F_C2) / 
                   (Fig3_E | Fig3_F) / 
                   (S4_B_D | S4_F_D) / 
                   (S4_G | plot_spacer()) +
  plot_annotation(tag_levels = 'A') &
  theme(plot.margin = margin(1,1,1,1))
Fig_S4_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Fig_S4_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig_S4_combined, 
       width = 10, height = 16, dpi = 600)

```


# Figure S5 - boxplots - plant variables

```{r}

library(ggpubr)
library(ggplot2)
library(rstatix)
library(dplyr)
library(stringr)

# Define the variable names you're interested in
start_col <- which(names(plant_data_dec23) == "Total Soluble Solids")
end_col <- which(names(plant_data_dec23) == "Total Fruit Hand Harvest Estimate")
target_vars <- c("Root Length Colonized", names(plant_data_dec23)[start_col:end_col])

# Create a list to store plots
plot_list <- list()

# Generate each plot and store in list
for (var in target_vars) {
  
  # Perform statistical test
  stat_test <- plant_data_dec23 %>%
    t_test(as.formula(paste0("`", var, "` ~ Treatment"))) %>%
    add_significance("p") %>%
    add_xy_position(x = "Treatment")
  
  # Format variable name: only capitalize first word
  var_label <- str_to_sentence(var)
  
  # Create base plot
  p <- ggboxplot(plant_data_dec23, 
                 x = "Treatment", 
                 y = var,
                 fill = "Treatment",
                 palette = amf_colors,
                 title = "",
                 xlab = "",
                 ylab = var_label) +
    theme_bw() +
    theme(legend.position = "none",
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank()) +
    stat_pvalue_manual(
      stat_test,
      label = "p.signif",
      tip.length = 0.02,
      hide.ns = FALSE,
      label.size = 8
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)))  # 5% bottom, 15% top

  
  plot_list[[var]] <- p
}

# Arrange all plots on one page
combined_plot_2 <- ggarrange(plotlist = plot_list, 
                             ncol = 3,
                             nrow = ceiling(length(plot_list) / 3),
                             common.legend = FALSE,
                             labels = "AUTO")  # Adds A, B, C, etc.

combined_plot_2

# Save to PDF
plot_file <- paste0(workdir, "plots/Manuscript_figures/S5_plant_boxplots_combined_", Sys.Date(), ".pdf")
ggsave(plot_file, combined_plot_2, width = 14, height = 10, device = "pdf")


```


# Desiccation test

Analysis of desiccation-related gene expression in rhizosphere metatranscriptomes
Based on desiccation-upregulated genes identified in Scales et al. 2023
DOI: 10.1111/1462-2920.16494

```{r}

library(dplyr)
library(stringr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(emmeans)

# ------------------------------------------------------------------------------
# Step 1: Define desiccation-upregulated genes from Scales et al. 2023 (Table 3)
# ------------------------------------------------------------------------------
# These genes were identified as significantly upregulated during desiccation
# stress in Curtobacterium (across Clades I, IV, and V)

desiccation_up_genes <- c(
  # Efflux pumps
  "macB",    # Antibiotic efflux pump
  "arsB",    # Arsenic efflux pump
  "cnrH",    # Efflux pump sigma factor
  
  # DNA repair and recombination
  "recA",    # DNA recombination
  "recN",    # DNA recombination
  "uvrA",    # DNA repair
  "dps",     # Negatively regulates DNA replication
  
  # SOS response
  "lexA",    # SOS Response
  "recX",    # SOS Response
  
  # Osmolyte synthesis (KEY desiccation markers)
  "dhaK",    # Osmolyte synthesis
  "dhaM",    # Osmolyte synthesis
  "treZ",    # Trehalose synthesis
  "treY",    # Trehalose synthesis
  "otsB",    # Trehalose synthesis
  
  # Oxidative stress response
  "katA",    # Reactive oxygen scavenging (catalase)
  "ahpF",    # Reactive oxygen scavenging
  
  # Sporulation and stress response
  "gdh",     # Sporulation
  "sigK",    # Sporulation sigma factor
  "rpoE",    # Stress response sigma factor
  
  # Other
  "tgnD",    # Degrades trigonelline
  "gabD",    # Degrades trigonelline
  "fadA"     # Fatty acid synthesis
)

# Create regex pattern for filtering
# This will match gene names in your Preferred_name column
gene_pattern <- paste0("\\b(", paste(desiccation_up_genes, collapse = "|"), ")\\b")

# ------------------------------------------------------------------------------
# Step 2: Filter metaT data to desiccation-upregulated genes only
# ------------------------------------------------------------------------------

desiccation_df <- annotated_expression_long %>%
  filter(
    str_detect(Preferred_name, regex(gene_pattern, ignore_case = TRUE))
  ) %>%
  mutate(
    # Categorize genes by functional role (based on Scales et al. Table 3)
    gene_category = case_when(
      Preferred_name %in% c("dhaK", "dhaM", "treZ", "treY", "otsB") ~ "Osmolyte synthesis",
      Preferred_name %in% c("recA", "recN", "uvrA", "dps") ~ "DNA repair/recombination",
      Preferred_name %in% c("katA", "ahpF") ~ "Oxidative stress",
      Preferred_name %in% c("lexA", "recX") ~ "SOS response",
      Preferred_name %in% c("macB", "arsB", "cnrH") ~ "Efflux pumps",
      Preferred_name %in% c("gdh", "sigK", "rpoE") ~ "Sporulation/stress sigma factors",
      TRUE ~ "Other desiccation-related"
    )
  )

# Print summary of genes found
cat("\n=== Genes found in dataset ===\n")
desiccation_df %>%
  count(Preferred_name, gene_category) %>%
  arrange(gene_category, Preferred_name) %>%
  print(n = 100)

cat("\nTotal unique genes found:", length(unique(desiccation_df$Preferred_name)), "/", length(desiccation_up_genes), "\n")

# ------------------------------------------------------------------------------
# Step 3: Calculate relative abundance of desiccation-upregulated genes
# ------------------------------------------------------------------------------

# Get total abundance per sample for normalization
total_by_sample <- annotated_expression_long %>%
  group_by(SampleID) %>%
  summarise(total_abundance = sum(abundance, na.rm = TRUE))

# Calculate relative abundance of desiccation genes
desiccation_rel <- desiccation_df %>%
  group_by(SampleID, Timepoint, Treatment, Site) %>%
  summarise(desiccation_abundance = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  left_join(total_by_sample, by = "SampleID") %>%
  mutate(rel_desiccation_abundance = desiccation_abundance / total_abundance)

# Convert timepoint to numeric weeks for analysis
desiccation_rel <- desiccation_rel %>%
  mutate(
    Timepoint = factor(Timepoint, levels = c("T1", "T2", "T3"), ordered = TRUE),
    week = case_when(
      Timepoint == "T1" ~ 9,
      Timepoint == "T2" ~ 14,
      Timepoint == "T3" ~ 19,
      TRUE ~ NA_real_
    )
  )

# ------------------------------------------------------------------------------
# Step 4: Statistical test - Linear Mixed Model
# ------------------------------------------------------------------------------
# Research question: Do control plants show NO increase in desiccation genes?
# H0: slope for Control = 0 (no temporal trend)
# H1: slope for Control ≠ 0

# Fit model with interaction to test treatment-specific slopes
model_interaction <- lmer(
  rel_desiccation_abundance ~ week * Treatment + (1|Site), 
  data = desiccation_rel
)

cat("\n=== Mixed Model Results ===\n")
summary(model_interaction)

# Extract coefficients for interpretation
model_summary <- summary(model_interaction)
coef_table <- coef(model_summary)

# Test specific hypothesis: is Control slope = 0?
control_slope <- coef_table["week", "Estimate"]
control_slope_se <- coef_table["week", "Std. Error"]
control_slope_p <- coef_table["week", "Pr(>|t|)"]

cat("\n=== Key Results ===\n")
cat("Control plants temporal trend (week effect):\n")
cat("  Slope:", round(control_slope, 6), "\n")
cat("  Std. Error:", round(control_slope_se, 6), "\n")
cat("  p-value:", round(control_slope_p, 4), "\n")
if (control_slope_p > 0.05) {
  cat("  Interpretation: No significant increase in desiccation genes in control plants (p > 0.05)\n")
} else {
  cat("  Interpretation: Significant temporal change detected (p < 0.05)\n")
}

# ------------------------------------------------------------------------------
# Step 5: Visualization
# ------------------------------------------------------------------------------

# Main plot: Temporal trend
desiccation_plot <- ggplot(desiccation_rel, 
                           aes(x = week, y = rel_desiccation_abundance, 
                               color = Treatment)) +
  geom_point(alpha = 0.5, size = 3) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_color_manual(
    breaks = c("Control", "Treatment"),
    values = c("#72b01d", "#B53471"),
    labels = c("Control", "Drought Treatment")
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    x = "Time (weeks)",
    y = "Relative abundance of desiccation-upregulated genes",
    color = "Treatment",
    title = "Expression of desiccation stress marker genes",
    subtitle = "Based on Scales et al. 2023 - genes upregulated during desiccation in Curtobacterium",
    caption = "Linear mixed model with site as random effect"
  )

print(desiccation_plot)

# Save plot
ggsave(
  paste0(workdir, "plots/metaT/desiccation_genes_Scales2023_", Sys.Date(), ".pdf"), 
  plot = desiccation_plot, 
  width = 10, height = 7, dpi = 300
)

#####################################
# ==============================================================================
# CODE TO GENERATE TABLE S3 - Complete version with all variable extraction
# ==============================================================================

# Your existing model
model_interaction <- lmer(
  rel_desiccation_abundance ~ week * Treatment + (1|Site), 
  data = desiccation_rel
)

# Get slopes for each treatment
emtrends_result <- emtrends(model_interaction, ~ Treatment, var = "week")

# Test if slopes are different from 0
slope_tests <- test(emtrends_result, null = 0)

# ------------------------------------------------------------------------------
# Step 2: Extract values for the table
# ------------------------------------------------------------------------------

# Get the summary data frames
slopes_summary <- summary(emtrends_result)
test_summary <- summary(slope_tests)

# Extract values for Control
control_slope <- slopes_summary$week.trend[slopes_summary$Treatment == "Control"]
control_SE <- slopes_summary$SE[slopes_summary$Treatment == "Control"]
control_df <- test_summary$df[test_summary$Treatment == "Control"]
control_p <- test_summary$p.value[test_summary$Treatment == "Control"]
control_t <- test_summary$t.ratio[test_summary$Treatment == "Control"]
control_CI_lower <- slopes_summary$lower.CL[slopes_summary$Treatment == "Control"]
control_CI_upper <- slopes_summary$upper.CL[slopes_summary$Treatment == "Control"]

# Extract values for Treatment (AMF)
treatment_slope <- slopes_summary$week.trend[slopes_summary$Treatment == "Treatment"]
treatment_SE <- slopes_summary$SE[slopes_summary$Treatment == "Treatment"]
treatment_df <- test_summary$df[test_summary$Treatment == "Treatment"]
treatment_p <- test_summary$p.value[test_summary$Treatment == "Treatment"]
treatment_t <- test_summary$t.ratio[test_summary$Treatment == "Treatment"]
treatment_CI_lower <- slopes_summary$lower.CL[slopes_summary$Treatment == "Treatment"]
treatment_CI_upper <- slopes_summary$upper.CL[slopes_summary$Treatment == "Treatment"]

# ------------------------------------------------------------------------------
# Step 3: Create Table S3A - Summary Statistics
# ------------------------------------------------------------------------------

table_S3A <- data.frame(
  Treatment = c("Control", "AMF-treated"),
  Slope = c(control_slope, treatment_slope),
  SE = c(control_SE, treatment_SE),
  CI_lower = c(control_CI_lower, treatment_CI_lower),
  CI_upper = c(control_CI_upper, treatment_CI_upper),
  t_value = c(control_t, treatment_t),
  df = c(control_df, treatment_df),
  p_value = c(control_p, treatment_p),
  Significance = c(
    ifelse(control_p < 0.05, "*", "ns"),
    ifelse(treatment_p < 0.05, "*", "ns")
  )
)

# Format for better display
table_S3A_formatted <- table_S3A %>%
  mutate(
    Slope_formatted = sprintf("%.2e", Slope),
    SE_formatted = sprintf("%.2e", SE),
    CI_95 = sprintf("[%.2e, %.2e]", CI_lower, CI_upper),
    t_formatted = sprintf("%.3f", t_value),
    df_formatted = sprintf("%.1f", df),
    p_formatted = sprintf("%.3f", p_value)
  ) %>%
  select(Treatment, Slope_formatted, SE_formatted, CI_95, t_formatted, 
         df_formatted, p_formatted, Significance)

# Rename columns for final table
colnames(table_S3A_formatted) <- c("Treatment", "Slope (week⁻¹)", "SE", 
                                   "95% CI", "t", "df", "p-value", "Sig.")

cat("\n=== TABLE S3A: Summary Statistics ===\n")
print(table_S3A_formatted, row.names = FALSE)

# Save table
write.csv(table_S3A_formatted, 
          paste0(workdir, "results/Table_S3A_desiccation_summary_", Sys.Date(), ".csv"),
          row.names = FALSE)

```
