---
title: "Tomato_manuscript_AMF"
author: "LJ"
date: "2026-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require("knitr")

opts_knit$set(root.dir = "/Users/ljiraska/Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/") 

```

# Load data

```{r include=FALSE}

# Load packages
library(microbiome)
library(phyloseq)
library(RColorBrewer)
library(ggpubr)
library(dplyr)
library(vegan)
library(ggplot2)
library(VennDiagram)
library(csv)
library(readr)
library(microViz)
library(ggConvexHull)
library(metagMisc)
library(ggrepel)

library(geodist)

library(eulerr)
library(ggpmisc)

library(patchwork)
library(pheatmap)

###
set.seed(10292023)

# Macbook
workdir = "/Users/ljiraska/Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/"

```


# Color scheme

```{r eval=FALSE, include=FALSE}
#### ============================================================
####  Color palette
#### ============================================================

### ggplot2 theme

theme_soil <- function(base_size = 16, legend = "none") {
  theme_bw(base_size = base_size) +
    theme(
      text = element_text(size = base_size),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = legend,
      strip.background = element_rect(fill = "grey90", color = "black"),
      strip.text = element_text(size = base_size * 0.9),
      axis.title = element_text(size = base_size * 1.1),
      axis.text = element_text(size = base_size * 0.9)
    )
}

### ------------------------------------------------------------
### 1. Soil compartment colors (bulk vs rhizosphere)
### ------------------------------------------------------------

compartment_colors <- c(
  "bulk"  = "#BA7F3A",  
  "rhizo" = "#3C8E3A"  
)


### ------------------------------------------------------------
### 2. Bulk & Rhizosphere across time (T0–T3)
### ------------------------------------------------------------

time_colors <- c(
  # Bulk (browns)
  "bulk_T0" = "#C9A46C",  
  "bulk_T1" = "#BA7F3A",
  "bulk_T2" = "#8F5A28",
  "bulk_T3" = "#5E4224",
  
  # Rhizosphere (greens)
  "rhizo_T1" = "#6BBF59",
  "rhizo_T2" = "#3C8E3A",
  "rhizo_T3" = "#1F5F2A"
)


### ------------------------------------------------------------
### 3. Bulk soil seasonal colors (month-level)
### ------------------------------------------------------------

bulk_month_colors <- c(
  "March"  = "#C9A46C",
  "June"   = "#BA7F3A",
  "July"   = "#8F5A28",
  "August" = "#5E4224"
)

### ------------------------------------------------------------
### 4. Rhizosphere seasonal colors (month-level)
### ------------------------------------------------------------

rhizo_month_colors <- c(
  "June" = "#6BBF59",
  "July" = "#3C8E3A",
  "August" = "#1F5F2A"
)

### ------------------------------------------------------------
### 5. Dry → Wet soil gradient
### ------------------------------------------------------------

moisture_gradient <- scale_color_gradient(
  low  = "#B8945E",   # darker tan (dry soil)
  high = "#08519C"    # darker blue (wet soil)
)

### ------------------------------------------------------------
### 6. Sampling plot colors
### ------------------------------------------------------------

plot_colors <- c(
  "1_4" = "#7B5FAE",  # deeper purple
  "5_5" = "#2D9CA2",  # strong teal
  "7_8" = "#E066A1"   # bright but not neon pink
)

### ------------------------------------------------------------
### 7. AMF treatment colors
### ------------------------------------------------------------

amf_colors <- c(
  "Control"   = "#B0B0B0",  # medium gray
  "Treatment" = "#344B7E"   # muted dark blue
)

direction_colors <- c(
    "Up"   = "#344B7E",
    "Down" = "#B0B0B0"
  )
### ------------------------------------------------------------
### 8. Heatmap colors
### ------------------------------------------------------------

heatmap_colors <- colorRampPalette(c("white", "#BBBBBB", "#888888", "#4D4D4D"))(200)

```

##############################################################
# Data Prep
##############################################################
# Upload data and create phyloseqs

Creates phyloseqs from 16S, ITS1, DNA viromes, RNA viromes

```{r}
##################################################################################################
### DNA Viromes

dnavir_otu <- read.csv(paste0(workdir,"./viromes/230427_covtab_norm.csv"))

dnavir_meta <- read.csv(paste0(workdir,"./viromes/220119_metadata.csv"))

## Check for metadata correctness

dnavir_meta_fix <- dnavir_meta %>% select(-sample, -soil_moisture_perc.1)

# save the correct metadata file 05/29/2024
write.csv(dnavir_meta_fix, paste0(workdir, "viromes/240529_metadata_LJ.csv"))

# OTU
str(dnavir_otu)
rownames(dnavir_otu) <- dnavir_otu$Contig
dnavir_otu <- dnavir_otu[,!(names(dnavir_otu) == "Contig")]
dim(dnavir_otu)

## Metadata file
rownames(dnavir_meta_fix) <- dnavir_meta_fix$Sample
colnames(dnavir_meta_fix)[colnames(dnavir_meta_fix) == "Sample"] <- "SampleID"
dim(dnavir_meta_fix)

physeq_DV <- phyloseq(otu_table(dnavir_otu, taxa_are_rows = TRUE), 
                     sample_data(dnavir_meta_fix))

physeq_DV

sort(phyloseq::sample_sums(physeq_DV))

saveRDS(physeq_DV, paste0(workdir,"./source_data/physeq_objects/physeq_DV_normalized_DNAViruses_MAY2023"))

###############################
### 16S
bacteria_otu <- read.csv(paste0(workdir,"amplicon/tomato_16S_norm.csv"))
bacteria_tax <- read_csv(paste0(workdir,"amplicon/tax1_16s.csv")) %>% as.data.frame()

bacteria_meta <- read.csv(paste0(workdir,"viromes/240529_metadata_LJ.csv"))

# OTU
str(bacteria_otu)
rownames(bacteria_otu) <- bacteria_otu$ASV
bacteria_otu <- bacteria_otu[,!(names(bacteria_otu) == "ASV")]
dim(bacteria_otu)

## Taxonomy file
str(bacteria_tax)
rownames(bacteria_tax) <- bacteria_tax$OTU_ID
bacteria_tax <- bacteria_tax[,!(names(bacteria_tax) == "OTU_ID")]
bacteria_tax.m <- as.matrix(bacteria_tax)

## Metadata file
rownames(bacteria_meta) <- bacteria_meta$X
bacteria_meta <- bacteria_meta[,-1]
dim(bacteria_meta)

physeq_B <- phyloseq(otu_table(bacteria_otu, taxa_are_rows = TRUE), 
                     tax_table(bacteria_tax.m), 
                     sample_data(bacteria_meta))

physeq_B

physeq_B3 = subset_taxa(physeq_B, !Family %in% c("Chloroplast","Mitochondria"))
physeq_B
physeq_B3

physeq_B3 = subset_taxa(physeq_B3, !Class %in% c("Chloroplast","Mitochondria"))
physeq_B
physeq_B3

physeq_B3 = subset_taxa(physeq_B3, !Order %in% c("Chloroplast","Mitochondria"))
physeq_B
physeq_B3

physeq_antiB3 = subset_taxa(physeq_B, Family %in% c("Chloroplast","Mitochondria") | Class %in% c("Chloroplast") | Order %in% c("Chloroplast"))

physeq_B3
physeq_antiB3

count_mitochondria <- sort(phyloseq::sample_sums(physeq_antiB3)) %>% sum()
count_NOmitochondria <- sort(phyloseq::sample_sums(physeq_B3)) %>% sum()

(count_mitochondria*100)/count_NOmitochondria

#### This above is inacurate as it is run on the normalized data
# We need the raw table
tomato_16S_raw <- read_csv("Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/amplicon/tomato_16S_raw.csv")

# Create vector vector_m_ch, from physeq_antiB3 that contains mitochondria and chloroplasts
vector_m_ch <- physeq_antiB3@tax_table %>% rownames() 

# Filter the rows where the first column matches vector_m_ch
otu_16S_raw_m_ch_filtr <- tomato_16S_raw %>% 
  filter(tomato_16S_raw[[1]] %in% vector_m_ch)

dim(otu_16S_raw_m_ch_filtr)

# Sum the reads for the filtered table (excluding the first column)
sum_filtered <- rowSums(otu_16S_raw_m_ch_filtr[ , -1]) %>% sum()

# Sum the reads for the whole table (excluding the first column)
sum_total <- rowSums(tomato_16S_raw[ , -1]) %>% sum()

# Calculate the percentage
percentage <- (sum_filtered / sum_total) * 100

# Output the result
percentage

physeq_B # 2049
physeq_B3 # 1979 taxa

sort(phyloseq::sample_sums(physeq_B3))

saveRDS(physeq_B, paste0(workdir,"./source_data/physeq_objects/physeq_B_normalized_withChlMit_MAY2023"))

saveRDS(physeq_B3, paste0(workdir,"./source_data/physeq_objects/physeq_B_normalized_noChlor_noMitoch_OCT2024"))

########################################################################
### ITS1
fungi_otu <- read.csv(paste0(workdir,"amplicon/tomato_ITS_norm.csv"))
fungi_tax <- read_csv(paste0(workdir,"amplicon/tax1_its.csv")) %>% as.data.frame()
fungi_meta <- read.csv(paste0(workdir,"viromes/240529_metadata_LJ.csv"))

# OTU
str(fungi_otu)
rownames(fungi_otu) <- fungi_otu$ASV
fungi_otu <- fungi_otu[,!(names(fungi_otu) == "ASV")]
dim(fungi_otu)

## Taxonomy file
str(fungi_tax)
rownames(fungi_tax) <- fungi_tax$OTU_ID
fungi_tax <- fungi_tax[,!(names(fungi_tax) == "OTU_ID")]
fungi_tax.m <- as.matrix(fungi_tax)

## Metadata file
rownames(fungi_meta) <- fungi_meta$SampleID
fungi_meta <- fungi_meta[,-1]
dim(fungi_meta)

physeq_F <- phyloseq(otu_table(fungi_otu, taxa_are_rows = TRUE), 
                     tax_table(fungi_tax.m), 
                     sample_data(fungi_meta))

physeq_F

# Taxonomic filtering
physeq_F1 <- subset_taxa(physeq_F, Kingdom =="k__Fungi")

physeq_F
physeq_F1

sort(phyloseq::sample_sums(physeq_F1))

saveRDS(physeq_F1, paste0(workdir,"./source_data/physeq_objects/physeq_F_normalized_Fungi_MAY2023"))

##################################################################################################
### RNA Viruses

rnavir_otu <- read.csv(paste0(workdir,"RNA/RNA_covtab_norm.csv"))
rnavir_meta <- read.csv(paste0(workdir,"viromes/220119_metadata.csv"))

## These files needs a bit of work before they can be input in phyloseq
# OTU
str(rnavir_otu)
rownames(rnavir_otu) <- rnavir_otu$Contig
rnavir_otu <- rnavir_otu[,!(names(rnavir_otu) == "Contig")]
dim(rnavir_otu)

## Metadata file
rownames(rnavir_meta) <- rnavir_meta$Sample
colnames(rnavir_meta)[colnames(rnavir_meta) == "Sample"] <- "SampleID"
dim(rnavir_meta)

rnavir_meta2 <- rnavir_meta %>% mutate(rhizo_rhizo_t = if_else(bulk_rhizo_t == "rhizo", paste0(bulk_rhizo, "_" , Timepoint, sep = " "), bulk_rhizo_t))

physeq_RV <- phyloseq(otu_table(rnavir_otu, taxa_are_rows = TRUE), 
                      sample_data(rnavir_meta2))

physeq_RV

#########
sort(phyloseq::sample_sums(physeq_RV))

saveRDS(physeq_RV, paste0(workdir,"./source_data/physeq_objects/physeq_RV_normalized_RNAViruses_MAY2023"))

```


```{r}
# Raw phyloseqs

physeq_B <- readRDS(paste0(workdir,"./source_data/physeq_objects/physeq_B_normalized_noChlor_noMitoch_OCT2024"))
physeq_F1 <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_F_normalized_Fungi_MAY2023"))

physeq_DV <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_DV_normalized_DNAViruses_MAY2023"))
physeq_RV <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_RV_normalized_RNAViruses_MAY2023"))

# Hellinger trasnformed physeqs

ps_B_tr <- microbiome::transform(physeq_B, 'hellinger')
ps_F_tr <- microbiome::transform(physeq_F1, 'hellinger')

ps_DV_tr <- microbiome::transform(physeq_DV, 'hellinger')
ps_RV_tr <- microbiome::transform(physeq_RV, 'hellinger')


saveRDS(ps_B_tr, paste0(workdir,"./source_data/physeq_objects/physeq_B_normalized_noChlor_noMitoch_OCT2024_hellinger"))

saveRDS(ps_F_tr, paste0(workdir,"./source_data/physeq_objects/physeq_F_normalized_Fungi_MAY2023_hellinger"))

saveRDS(ps_DV_tr, paste0(workdir,"./source_data/physeq_objects/physeq_DV_normalized_DNAViruses_MAY2023_hellinger"))
saveRDS(ps_RV_tr, paste0(workdir,"./source_data/physeq_objects/physeq_RV_normalized_RNAViruses_MAY2023_hellinger"))

# Read Hellinger transformed physeqs only
ps_B_tr <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_B_normalized_noChlor_noMitoch_OCT2024_hellinger"))
ps_F_tr <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_F_normalized_Fungi_MAY2023_hellinger"))

ps_DV_tr <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_DV_normalized_DNAViruses_MAY2023_hellinger"))
ps_RV_tr <- readRDS(paste0(workdir, "source_data/physeq_objects/physeq_RV_normalized_RNAViruses_MAY2023_hellinger"))

```


##############################################################
# Analysis
##############################################################


# Figure 1

## Panel A

This is a donut plot showing the number of vOTUs from different ecosystems in PIGEON, without vOTUs from Russel Range.

```{r}
library(readr)
df_PIGEON_donut_plot <- read_csv("Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/source_data/pigeon/df_PIGEON_donut_plot.csv")


############
habitat_colors_donut <- c(
  "Marine" = "#08306B",                        # dark blue
  "Freshwater" = "#6BAED6",                    # light blue
  "Non-marine Saline and Alkaline" = "#41B6C4", # teal
  "Rhizosphere Soil" = "#3C8E3A",               # rhizo green (consistent!)
  "Peat soil" = "#4B5D2A",                      # peat olive
  "Natural Soil" = "#C7A76C",                   # light brown
  "Agricultural" = "#8C2D19",                   # reddish brown
  "Digestive" = "#9E9AC8",                      # muted purple
  "Other" = "#BDBDBD"                           # grey
)


library(tidyverse)

Fig1_A <- df_PIGEON_donut_plot %>%
  arrange(desc(percentage)) %>%
  mutate(
    Habitat_type = factor(Habitat_type, levels = Habitat_type)
  ) %>%
  ggplot(aes(x = 2, y = percentage, fill = Habitat_type)) +
  geom_col(color = "white", width = 1) +
  coord_polar(theta = "y", start = pi / 2) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = habitat_colors_donut) +
  theme_void() +
  labs(fill = "Habitat type") + annotate(
    "text",
    x = 0,
    y = 0,
    label = scales::comma(sum(df_PIGEON_donut_plot$total_vOTUs)),
    size = 5,
    fontface = "bold"
  )+ geom_text(
  aes(label = paste0(sprintf("%.1f", percentage), "%")),
  position = position_stack(vjust = 0.5),
  size = 8,
  color = "white"
)

# Bigger legend text

Fig1_A <- df_PIGEON_donut_plot %>%
  arrange(desc(percentage)) %>%
  mutate(
    Habitat_type = factor(Habitat_type, levels = Habitat_type)
  ) %>%
  ggplot(aes(x = 2, y = percentage, fill = Habitat_type)) +
  geom_col(color = "white", width = 1) +
  coord_polar(theta = "y", start = pi / 2) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = habitat_colors_donut) +
  theme_void() +
  labs(fill = "Habitat type") +
  geom_text(
    aes(label = paste0(sprintf("%.1f", percentage), "%")),
    position = position_stack(vjust = 0.5),
    size = 8,
    color = "white"
  ) +
  theme(
    legend.text  = element_text(size = 18),
    legend.title = element_text(size = 19)
  )


Fig1_A

ggsave(
  Fig1_A,
  filename = paste0(workdir,"/plots/pigeon/Fig1_A_",Sys.Date(), ".pdf"),
  width = 10, height = 6, dpi = 600,
)

```


## Panel B

Map

```{r}

# Load coordinates data for PIGEON mapping
library(readxl)
coordinates_pigeon <- read_excel("Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/source_data/pigeon/df_PIGEON_mapping_coordinates.xlsx")

# Filter out all Russel Ranch vOTUs and no coordinates reads.

library(dplyr)

df_map_clean <- coordinates_pigeon_clean %>%
  # convert lat/lon to numeric (non-convertible -> NA)
  mutate(
    lat = as.numeric(lat),
    lon = as.numeric(lon)
  ) %>%
  # filter out bad rows
  filter(
    !is.na(lat),
    !is.na(lon),
    RR != "yes",
    !is.na(location),
    !location %in% c("no", "no location"),
    !str_detect(location, regex("multiple locations|not sure", ignore_case = TRUE))
  )

dim(coordinates_pigeon_clean)
dim(df_map_clean)

df_map <- df_map_clean %>%
  group_by(location, habitat_type, lat, lon) %>%
  summarise(
    n_vOTUs = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

###########################
library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

world <- ne_countries(scale = "medium", returnclass = "sf")

# make sure df_map is numeric and habitat_type is factor
df_map <- df_map %>%
  mutate(
    lat = as.numeric(lat),
    lon = as.numeric(lon),
    habitat_type = factor(habitat_type)
  )

df_map <- df_map %>%
  mutate(
    habitat_type = case_when(
      habitat_type %in% c("rhizosphere_soil") ~ "Rhizosphere Soil",
      habitat_type %in% c("agricultural_soil") ~ "Agricultural",
      habitat_type %in% c("natural soil", "natural_soil") ~ "Natural Soil",
      habitat_type %in% c("peat") ~ "Peat soil",
      habitat_type %in% c("aquatic") ~ "Freshwater", # or Marine if appropriate
      habitat_type %in% c("other") ~ "Other",
      TRUE ~ habitat_type  # leave others as-is
    )
  )

df_map$habitat_type <- factor(df_map$habitat_type,
                              levels = names(habitat_colors_donut))

df_map_world <- df_map %>%
  filter(!(lat >= 32 & lat <= 42 & lon >= -125 & lon <= -114))


# World map
Fig1_B_W <- ggplot() +
  geom_sf(data = world, fill = "grey95", color = "grey70", linewidth = 0.2) +
  geom_point(
    data = df_map_world,
    aes(x = lon, y = lat, size = n_vOTUs, fill = habitat_type),
    shape = 21,
    color = "black",
    alpha = 0.75
  ) +
  # size scaling: sqrt transform so small counts are bigger visually
  scale_size_area(
    max_size = 8,
    name = "Number of vOTUs",
    breaks = c(1, 2, 5, 10),  # optional nice legend
    trans = "sqrt"
  ) +
  scale_fill_manual(values = habitat_colors_donut) +
  coord_sf(expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Sample type"
  )

# Or as PDF for vector graphics
ggsave(
  Fig1_B_W,
  filename = paste0(workdir,"/plots/pigeon/Fig1_B_W_PIGEON_mapping_world_vOTU_map", Sys.Date(), ".pdf"),
  width = 7, height = 4, dpi = 600,
)

```

California only

```{r}
########## California only
# Filter to California only

df_map_CA <- df_map %>%
  filter((lat >= 32 & lat <= 42 & lon >= -125 & lon <= -114))

#########

library(tidyverse)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

# Load US states
us_states <- ne_states(country = "United States of America",
                       returnclass = "sf")

# Extract California
california <- us_states %>%
  filter(name == "California")

df_CA <- df %>%
  filter(country == "USA",
         lat > 32, lat < 42,
         lon > -125, lon < -114)

Fig1_B_CA <- ggplot() +
  geom_sf(
    data = california,
    fill = "grey95",
    color = "black",
    linewidth = 0.4
  ) +
  geom_point(
    data = df_map_CA,
    aes(x = lon, y = lat, size = n_vOTUs, fill = habitat_type),
    shape = 21,
    color = "black",
    alpha = 0.6,
    position = position_jitter(width = 0.08, height = 0.08)
  ) +
  scale_size_area(
    max_size = 20,
    name = "Number of vOTUs",
    trans = "sqrt",
    breaks = c(1, 10, 100, 1000)
  ) +
  scale_fill_manual(values = habitat_colors_donut) +
  coord_sf(
    xlim = st_bbox(california)[c("xmin", "xmax")],
    ylim = st_bbox(california)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Sample type",
    title = "California vOTU Sampling Sites"
  )


ggsave(
  Fig1_B_CA,
  filename = paste0(workdir,"/plots/pigeon/Fig1_B_CA_PIGEON_mapping_CA_vOTU_map_", Sys.Date(), ".pdf"),
  width = 7, height = 9, dpi = 600,
)


##### Modification

df_map_CA_modified <- df_map_CA %>%
  mutate(vOTU_bin = cut(
    n_vOTUs,
    breaks = c(0, 10, 50, 100, 500, 1000, Inf),
    labels = c("<10", "10–50", "50–100", "100–500", "500–1000", ">1000"),
    include.lowest = TRUE,
    right = TRUE
  ))

# force levels order to match numeric progression
df_map_CA_modified$vOTU_bin <- factor(
  df_map_CA_modified$vOTU_bin,
  levels = c("<10", "10–50", "50–100", "100–500", "500–1000", ">1000")
)

size_values <- c(3, 4, 5, 7, 10, 14)
names(size_values) <- levels(df_map_CA_modified$vOTU_bin)

Fig1_B_CA_modified <- ggplot() +
  geom_sf(data = california, fill = "grey95", color = "black") +
  geom_point(
    data = df_map_CA_modified,
    aes(x = lon, y = lat, size = vOTU_bin, fill = habitat_type),
    shape = 21,
    color = "black",
    alpha = 0.8,
    position = position_jitter(width = 0.1, height = 0.1)
  ) +
  scale_size_manual(values = size_values, name = "Number of vOTUs") +
  scale_fill_manual(values = habitat_colors_donut) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Sample type",
    title = "California vOTU Sampling Sites"
  )


ggsave(
  Fig1_B_CA_modified,
  filename = paste0(workdir,"/plots/pigeon/Fig1_B_CA_modified_PIGEON_mapping_CA_vOTU_map_", Sys.Date(), ".pdf"),
  width = 10, height = 15, dpi = 600,
)

```


## Panel C

```{r}
library(readr)
df_PIGEON_mapping_donut_plot <- read_csv("Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/source_data/pigeon/df_PIGEON_mapping_donut_plot.csv")

############
habitat_colors_donut <- c(
  "Marine" = "#08306B",                        # dark blue
  "Freshwater" = "#6BAED6",                    # light blue
  "Non-marine Saline and Alkaline" = "#41B6C4", # teal
  "Rhizosphere Soil" = "#3C8E3A",               # rhizo green (consistent!)
  "Peat soil" = "#4B5D2A",                      # peat olive
  "Natural Soil" = "#C7A76C",                   # light brown
  "Agricultural" = "#8C2D19",                   # reddish brown
  "Digestive" = "#9E9AC8",                      # muted purple
  "Other" = "#BDBDBD"                           # grey
)


library(tidyverse)
Fig1_C <- df_PIGEON_mapping_donut_plot %>%
  filter(percentage > 0) %>% 
  arrange(desc(percentage)) %>%
  mutate(
    Habitat_type = factor(Habitat_type, levels = Habitat_type)
  ) %>%
  ggplot(aes(x = 2, y = percentage, fill = Habitat_type)) +
  geom_col(color = "white", width = 1) +
  coord_polar(theta = "y", start = pi / 2) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = habitat_colors_donut) +
  theme_void() +
  labs(fill = "Habitat type") + annotate(
    "text",
    x = 0,
    y = 0,
    label = scales::comma(sum(df_PIGEON_donut_plot$total_vOTUs)),
    size = 5,
    fontface = "bold"
  )+ geom_text(
    aes(label = paste0(sprintf("%.1f", percentage), "%")),
    position = position_stack(vjust = 0.5),
    size = 8,
    color = "white"
  )

Fig1_C <- df_PIGEON_mapping_donut_plot %>%
  filter(percentage > 0) %>%
  arrange(desc(percentage)) %>%
  mutate(
    Habitat_type = factor(Habitat_type, levels = Habitat_type)
  ) %>%
  ggplot(aes(x = 2, y = percentage, fill = Habitat_type)) +
  geom_col(color = "white", width = 1) +
  coord_polar(theta = "y", start = pi / 2) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = habitat_colors_donut) +
  theme_void() +
  labs(fill = "Habitat type") +
  geom_text(
    aes(label = paste0(sprintf("%.1f", percentage), "%")),
    position = position_stack(vjust = 0.5),
    size = 8,
    color = "white"
  ) +
  theme(
    legend.text  = element_text(size = 18),
    legend.title = element_text(size = 19)
  )


Fig1_C

ggsave(
  Fig1_C,
  filename = paste0(workdir,"/plots/pigeon/Fig1_C_",Sys.Date(), ".pdf"),
  width = 10, height = 6, dpi = 600,
)


# Merged Fig1C data

df_PIGEON_mapping_donut_merged <- df_PIGEON_mapping_donut_plot %>%
  mutate(
    Habitat_type = ifelse(percentage < 2,
                          "Other",
                          Habitat_type)
  ) %>%
  group_by(Habitat_type) %>%
  summarise(
    total_vOTUs = sum(total_vOTUs),
    percentage  = sum(percentage),
    .groups = "drop"
  )


Fig1_C_merged <- df_PIGEON_mapping_donut_merged %>%
  arrange(desc(percentage)) %>%
  mutate(
    Habitat_type = factor(Habitat_type, levels = Habitat_type)
  ) %>%
  ggplot(aes(x = 2, y = percentage, fill = Habitat_type)) +
  geom_col(color = "white", width = 1) +
  coord_polar(theta = "y", start = pi / 2) +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = habitat_colors_donut) +
  theme_void() +
  labs(fill = "Habitat type") +
  geom_text(
    aes(label = paste0(sprintf("%.1f", percentage), "%")),
    position = position_stack(vjust = 0.5),
    size = 8,
    color = "white"
  ) +
  theme(
    legend.text  = element_text(size = 18),
    legend.title = element_text(size = 19)
  )


Fig1_C_merged

ggsave(
  Fig1_C_merged,
  filename = paste0(workdir,"/plots/pigeon/Fig1_C_merged_",Sys.Date(), ".pdf"),
  width = 10, height = 6, dpi = 600,
)
```

## Combined Figure 1

```{r}
#Combine the two maps

# Combine with inset
combined_Fig1_maps <- ggdraw() +
  draw_plot(Fig1_B_CA_modified + theme(legend.position = "none")) +
  draw_plot(Fig1_B_W + theme(legend.position = "none"), x = 0.40, y = 0.5, width = 0.6, height = 0.6)

```

```{r}

library(patchwork)

# Remove legends from donuts
Fig1_A_no_legend <- Fig1_A + theme(legend.position = "none")
Fig1_C_no_legend <- Fig1_C_merged + theme(legend.position = "none")

# Stack the donuts vertically
donuts_stack <- Fig1_A_no_legend / Fig1_C_no_legend 

combined_Fig1_maps_trimmed <- combined_Fig1_maps +
  coord_sf(expand = FALSE) +
  theme(legend.position = "none")  # already removing legend

# Combine with California map side by side
Figure_1_combined <- donuts_stack | combined_Fig1_maps_trimmed +
  plot_annotation(tag_levels = 'A')

Figure_1_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_1_combined_2_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_1_combined, 
       width = 15, height = 10, dpi = 600)

```


# Figure 2

## Panel A

```{r}
###########################################################################################
# transform the data
ps_DV_tr <- microbiome::transform(physeq_DV, 'hellinger')

# PCoA - with Bray Curtis
pcoa_DV_tr <- ordinate(ps_DV_tr, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_DV_tr <- as.data.frame(pcoa_DV_tr$vectors)

# Extract sample data for temporal information
temporal_data_DV_tr <- as.data.frame(sample_data(ps_DV_tr))

# Combine data for plotting
combined_data_DV_tr <- cbind(temporal_data_DV_tr, pcoa_result_DV_tr)

# Calculate centroids for each group and timepoint
centroid_data_DV_tr <- combined_data_DV_tr %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
combined_data_DV_tr$Timepoint <- factor(combined_data_DV_tr$Timepoint, levels = c("T0", "T1", "T2", "T3"))

# Arrange data for plotting
arranged_data_DV_tr <- centroid_data_DV_tr %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_DV_tr)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_DV_tr)$bulk_rhizo, 
                                                "_", sample_data(ps_DV_tr)$Timepoint)

# Panel A
Fig2_A <- ggplot(combined_data_DV_tr, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo_t_2), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr$values$Eigenvalues[1] / sum(pcoa_DV_tr$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr$values$Eigenvalues[2] / sum(pcoa_DV_tr$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

ggsave(paste0(workdir, "plots/pcoa/trajectory_pcoa_DV_tr_DIFF_colors_hull_NO_LEGEND_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_A,
       width = 7, height = 6, dpi = 600)


### Legend only

Fig2_legend <- plot_ordination(ps_DV_tr, pcoa_DV_tr, type = "samples", color = "bulk_rhizo_t_2", shape = "bulk_rhizo") + 
  geom_point(size = 6, alpha=1) +
  scale_color_manual(
    name = 'Time', 
    values = time_colors
  ) +
  geom_convexhull(aes(colour=as.character(bulk_rhizo)), alpha = 0.05) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr$values$Eigenvalues[1] / sum(pcoa_DV_tr$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr$values$Eigenvalues[2] / sum(pcoa_DV_tr$values$Eigenvalues) * 100, 2), "%)")) + theme_bw()

ggsave(paste0(workdir, "plots/pcoa/Fig2_legend_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_legend,
       width = 7, height = 6, dpi = 600)

```

## Panel B

####### Version 3

```{r}

# PCoA - with Bray Curtis

### T1
ps_DV_tr_T1 <- ps_filter(ps_DV_tr, Timepoint == "T1")

pcoa_DV_tr_T1 <- ordinate(ps_DV_tr_T1, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_DV_tr_T1 <- as.data.frame(pcoa_DV_tr_T1$vectors)

# Extract sample data for temporal information
temporal_data_DV_tr_T1 <- as.data.frame(sample_data(ps_DV_tr_T1))

# Combine data for plotting
combined_data_DV_tr_T1 <- cbind(temporal_data_DV_tr_T1, pcoa_result_DV_tr_T1)

# Calculate centroids for each group and timepoint
centroid_data_DV_tr_T1 <- combined_data_DV_tr_T1 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
combined_data_DV_tr_T1$Timepoint <- factor(combined_data_DV_tr_T1$Timepoint, levels = c("T0", "T1", "T2", "T3"))

# Arrange data for plotting
arranged_data_DV_tr_T1 <- centroid_data_DV_tr_T1 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_DV_tr_T1)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_DV_tr_T1)$bulk_rhizo, 
                                                "_", sample_data(ps_DV_tr_T1)$Timepoint)

# Panel A

Fig2_B_T1_3 <- ggplot(combined_data_DV_tr_T1, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_T1$values$Eigenvalues[1] / sum(pcoa_DV_tr_T1$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_T1$values$Eigenvalues[2] / sum(pcoa_DV_tr_T1$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_Fig2_B_T1_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T1_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/Fig2_B_T1_3_DV_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T1_3,
       width = 4, height = 3, dpi = 600)

### T2

ps_DV_tr_T2 <- ps_filter(ps_DV_tr, Timepoint == "T2")

pcoa_DV_tr_T2 <- ordinate(ps_DV_tr_T2, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_DV_tr_T2 <- as.data.frame(pcoa_DV_tr_T2$vectors)

# Extract sample data for temporal information
temporal_data_DV_tr_T2 <- as.data.frame(sample_data(ps_DV_tr_T2))

# Combine data for plotting
combined_data_DV_tr_T2 <- cbind(temporal_data_DV_tr_T2, pcoa_result_DV_tr_T2)

# Calculate centroids for each group and timepoint
centroid_data_DV_tr_T2 <- combined_data_DV_tr_T2 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
#combined_data_DV_tr_T2$Timepoint <- factor(combined_data_DV_tr_T2$Timepoint, levels = c("T0", "T2", "T2", "T3"))

# Arrange data for plotting
arranged_data_DV_tr_T2 <- centroid_data_DV_tr_T2 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_DV_tr_T2)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_DV_tr_T2)$bulk_rhizo, 
                                                   "_", sample_data(ps_DV_tr_T2)$Timepoint)

# Panel A
Fig2_B_T2_3 <- ggplot(combined_data_DV_tr_T2, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_T2$values$Eigenvalues[1] / sum(pcoa_DV_tr_T2$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_T2$values$Eigenvalues[2] / sum(pcoa_DV_tr_T2$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_Fig2_B_T2_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T2_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/Fig2_B_T2_3_DV_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T2_3,
       width = 4, height = 3, dpi = 600)

### T3

ps_DV_tr_T3 <- ps_filter(ps_DV_tr, Timepoint == "T3")

pcoa_DV_tr_T3 <- ordinate(ps_DV_tr_T3, "PCoA", distance = "bray", binary = F)

# Extract PCoA results
pcoa_result_DV_tr_T3 <- as.data.frame(pcoa_DV_tr_T3$vectors)

# Extract sample data for temporal information
temporal_data_DV_tr_T3 <- as.data.frame(sample_data(ps_DV_tr_T3))

# Combine data for plotting
combined_data_DV_tr_T3 <- cbind(temporal_data_DV_tr_T3, pcoa_result_DV_tr_T3)

# Calculate centroids for each group and timepoint
centroid_data_DV_tr_T3 <- combined_data_DV_tr_T3 %>%
  group_by(bulk_rhizo, Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

# Convert Timepoint to factor with the desired order
#combined_data_DV_tr_T3$Timepoint <- factor(combined_data_DV_tr_T3$Timepoint, levels = c("T0", "T3", "T3", "T3"))

# Arrange data for plotting
arranged_data_DV_tr_T3 <- centroid_data_DV_tr_T3 %>%
  arrange(bulk_rhizo, Timepoint)

sample_data(ps_DV_tr_T3)$bulk_rhizo_t_2 <-  paste0(sample_data(ps_DV_tr_T3)$bulk_rhizo, 
                                                   "_", sample_data(ps_DV_tr_T3)$Timepoint)

# Panel A
Fig2_B_T3_3 <- ggplot(combined_data_DV_tr_T3, aes(Axis.1, Axis.2)) +
  geom_point(aes(shape = bulk_rhizo, fill = bulk_rhizo), size = 6, color = "black") +
  geom_convexhull(aes(group = bulk_rhizo, fill = bulk_rhizo, color = bulk_rhizo), alpha = 0.1) +
  scale_shape_manual(values = c(21, 24)) +
  scale_fill_manual(values = c(compartment_colors, time_colors)) +
  scale_color_manual(values = c(compartment_colors)) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_T3$values$Eigenvalues[1] / sum(pcoa_DV_tr_T3$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_T3$values$Eigenvalues[2] / sum(pcoa_DV_tr_T3$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()

ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_Fig2_B_T3_3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T3_3,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/Fig2_B_T3_3_DV_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T3_3,
       width = 4, height = 3, dpi = 600)

```


```{r}
#### T1
# Subset to T1 only
ps_DV_tr_T1 <- ps_filter(ps_DV_tr, Timepoint == "T1")

# Split by compartment
ps_DV_tr_T1_bulk  <- ps_filter(ps_DV_tr_T1, bulk_rhizo == "bulk")
ps_DV_tr_T1_rhizo <- ps_filter(ps_DV_tr_T1, bulk_rhizo == "rhizo")

# Extract taxa
bulk_taxa_DV_T1 <- taxa_names(ps_DV_tr_T1_bulk)[taxa_sums(ps_DV_tr_T1_bulk) > 0]

rhizo_taxa_DV_T1 <- taxa_names(ps_DV_tr_T1_rhizo)[taxa_sums(ps_DV_tr_T1_rhizo) > 0]

# Prepare Eulerr

euler_input_DV_T1 <- list(Bulk = bulk_taxa_DV_T1, 
                          Rhizosphere = rhizo_taxa_DV_T1)

# Plot

euler_fit_DV_T1 <- euler(euler_input_DV_T1)


Fig2_B_T1_3_eulerr <- plot(
  euler_fit_DV_T1,
  fills = c("#BA7F3A", "#6BBF59"),
  edges = TRUE,
  quantities = TRUE,
  main = "T1"
)

ggsave(paste0(workdir, "plots/euler/Fig2_B_T1_3_eulerr_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T1_3_eulerr,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/euler/Fig2_B_T1_3_eulerr_DV_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T1_3_eulerr,
       width = 4, height = 3, dpi = 600)

### T2

# Subset to T2 only
ps_DV_tr_T2 <- ps_filter(ps_DV_tr, Timepoint == "T2")

# Split by compartment
ps_DV_tr_T2_bulk  <- ps_filter(ps_DV_tr_T2, bulk_rhizo == "bulk")
ps_DV_tr_T2_rhizo <- ps_filter(ps_DV_tr_T2, bulk_rhizo == "rhizo")

# Extract taxa
bulk_taxa_DV_T2 <- taxa_names(ps_DV_tr_T2_bulk)[taxa_sums(ps_DV_tr_T2_bulk) > 0]

rhizo_taxa_DV_T2 <- taxa_names(ps_DV_tr_T2_rhizo)[taxa_sums(ps_DV_tr_T2_rhizo) > 0]

# Prepare Eulerr

euler_input_DV_T2 <- list(Bulk = bulk_taxa_DV_T2, 
                          Rhizosphere = rhizo_taxa_DV_T2)


# Plot

euler_fit_DV_T2 <- euler(euler_input_DV_T2)


Fig2_B_T2_3_eulerr <- plot(
  euler_fit_DV_T2,
  fills = c("#8F5A28", "#3C8E3A"),
  edges = TRUE,
  quantities = TRUE,
  main = "T2"
)

ggsave(paste0(workdir, "plots/euler/Fig2_B_T2_3_eulerr_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T2_3_eulerr,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/euler/Fig2_B_T2_3_eulerr_DV_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T2_3_eulerr,
       width = 4, height = 3, dpi = 600)

### T3

# Subset to T3 only
ps_DV_tr_T3 <- ps_filter(ps_DV_tr, Timepoint == "T3")

# Split by compartment
ps_DV_tr_T3_bulk  <- ps_filter(ps_DV_tr_T3, bulk_rhizo == "bulk")
ps_DV_tr_T3_rhizo <- ps_filter(ps_DV_tr_T3, bulk_rhizo == "rhizo")

# Extract taxa
bulk_taxa_DV_T3 <- taxa_names(ps_DV_tr_T3_bulk)[taxa_sums(ps_DV_tr_T3_bulk) > 0]

rhizo_taxa_DV_T3 <- taxa_names(ps_DV_tr_T3_rhizo)[taxa_sums(ps_DV_tr_T3_rhizo) > 0]

# Prepare Eulerr

euler_input_DV_T3 <- list(Bulk = bulk_taxa_DV_T3, 
                    Rhizosphere = rhizo_taxa_DV_T3)


# Plot

euler_fit_DV_T3 <- euler(euler_input_DV_T3)


Fig2_B_T3_3_eulerr <- plot(
  euler_fit_DV_T3,
  fills = c("#5E4224", "#1F5F2A"),
  edges = TRUE,
  quantities = TRUE,
  main = "T3"
)

ggsave(paste0(workdir, "plots/euler/Fig2_B_T3_3_eulerr_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T3_3_eulerr,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/euler/Fig2_B_T3_3_eulerr_DV_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T3_3_eulerr,
       width = 4, height = 3, dpi = 600)

```

####### Version 2

```{r}

### Timepoint T1

ps_T1 <- subset_samples(physeq_DV, Timepoint == "T1")
ps_T1 <- prune_taxa(taxa_sums(ps_T1) > 0, ps_T1)

ps_T1_tr <- microbiome::transform(ps_T1, "hellinger")

pcoa_T1 <- ordinate(ps_T1_tr, "PCoA", distance = "bray", binary = FALSE)

scores_T1 <- as.data.frame(pcoa_T1$vectors)
meta_T1   <- as.data.frame(sample_data(ps_T1_tr))
plot_T1   <- cbind(meta_T1, scores_T1)

var1_T1 <- round(pcoa_T1$values$Eigenvalues[1] /
                   sum(pcoa_T1$values$Eigenvalues) * 100, 2)
var2_T1 <- round(pcoa_T1$values$Eigenvalues[2] /
                   sum(pcoa_T1$values$Eigenvalues) * 100, 2)

### Timepoint T2
ps_T2 <- subset_samples(physeq_DV, Timepoint == "T2")
ps_T2 <- prune_taxa(taxa_sums(ps_T2) > 0, ps_T2)

ps_T2_tr <- microbiome::transform(ps_T2, "hellinger")

pcoa_T2 <- ordinate(ps_T2_tr, "PCoA", distance = "bray", binary = FALSE)

scores_T2 <- as.data.frame(pcoa_T2$vectors)
meta_T2   <- as.data.frame(sample_data(ps_T2_tr))
plot_T2   <- cbind(meta_T2, scores_T2)

var1_T2 <- round(pcoa_T2$values$Eigenvalues[1] /
                   sum(pcoa_T2$values$Eigenvalues) * 100, 2)
var2_T2 <- round(pcoa_T2$values$Eigenvalues[2] /
                   sum(pcoa_T2$values$Eigenvalues) * 100, 2)


## Timepoint T3
ps_T3 <- subset_samples(physeq_DV, Timepoint == "T3")
ps_T3 <- prune_taxa(taxa_sums(ps_T3) > 0, ps_T3)

ps_T3_tr <- microbiome::transform(ps_T3, "hellinger")

pcoa_T3 <- ordinate(ps_T3_tr, "PCoA", distance = "bray", binary = FALSE)

scores_T3 <- as.data.frame(pcoa_T3$vectors)
meta_T3   <- as.data.frame(sample_data(ps_T3_tr))
plot_T3   <- cbind(meta_T3, scores_T3)

var1_T3 <- round(pcoa_T3$values$Eigenvalues[1] /
                   sum(pcoa_T3$values$Eigenvalues) * 100, 2)
var2_T3 <- round(pcoa_T3$values$Eigenvalues[2] /
                   sum(pcoa_T3$values$Eigenvalues) * 100, 2)

#############
Fig2_B_T1 <- ggplot(plot_T1,
                    aes(Axis.1, Axis.2,
                        color = bulk_rhizo,
                        shape = bulk_rhizo)) +
  geom_point(size = 6) +
  geom_convexhull(aes(fill = bulk_rhizo), alpha = 0.1) +
  scale_color_manual(values = c('#BA7F3A', '#3C8E3A')) +
  xlab(paste0("PCoA axis 1 (", var1_T1, "%)")) +
  ylab(paste0("PCoA axis 2 (", var2_T1, "%)")) +
  ggtitle("T1") +
  theme_soil()

Fig2_B_T1


Fig2_B_T2 <- ggplot(plot_T2,
                    aes(Axis.1, Axis.2,
                        color = bulk_rhizo,
                        shape = bulk_rhizo)) +
  geom_point(size = 6) +
  geom_convexhull(aes(fill = bulk_rhizo), alpha = 0.1) +
  scale_color_manual(values = c('#BA7F3A', '#3C8E3A')) +
  xlab(paste0("PCoA axis 1 (", var1_T2, "%)")) +
  ylab(paste0("PCoA axis 2 (", var2_T2, "%)")) +
  ggtitle("T2") +
  theme_soil()

Fig2_B_T2



Fig2_B_T3 <- ggplot(plot_T3,
                    aes(Axis.1, Axis.2,
                        color = bulk_rhizo,
                        shape = bulk_rhizo)) +
  geom_point(size = 6) +
  geom_convexhull(aes(fill = bulk_rhizo), alpha = 0.1) +
  scale_color_manual(values = c('#BA7F3A', '#3C8E3A')) +
  xlab(paste0("PCoA axis 1 (", var1_T3, "%)")) +
  ylab(paste0("PCoA axis 2 (", var2_T3, "%)")) +
  ggtitle("T3") +
  theme_soil()

Fig2_B_T3

############

ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_Fig2_B_T1_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T1,
       width = 7, height = 6, dpi = 600)


ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_Fig2_B_T2_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T2,
       width = 7, height = 6, dpi = 600)


ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_Fig2_B_T3_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B_T3,
       width = 7, height = 6, dpi = 600)
```


####### Previous

```{r}

Fig2_B_1 <- ggplot(combined_data_DV_tr, aes(x = Axis.1, y = Axis.2, color = bulk_rhizo, shape = bulk_rhizo)) +
  geom_point(size = 6, alpha = 1) +
  geom_convexhull(aes(fill = bulk_rhizo), alpha = 0.1) +
  scale_color_manual(values = c('#BA7F3A', '#3C8E3A')) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr$values$Eigenvalues[1] / sum(pcoa_DV_tr$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr$values$Eigenvalues[2] / sum(pcoa_DV_tr$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil() +
  facet_wrap(~Timepoint)

ggsave(paste0(workdir, "plots/pcoa/timepoints_pcoa_DV_tr_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_B,
       width = 7, height = 6, dpi = 600)

```

## Panel C

```{r}

# Plot the number of vOTUs between habitats and bulk soil moisture
alpha_DV_tr <- alpha(ps_DV_tr, index = "Observed")

alpha_DV_tr$SampleID <- rownames(alpha_DV_tr)
alpha_DV_tr_m <- merge(alpha_DV_tr, meta(ps_DV_tr), by = "SampleID")

alpha_DV_tr_m$Numeric_Timepoint <- case_when(
  alpha_DV_tr_m$Timepoint == "T0" ~ 0,
  alpha_DV_tr_m$Timepoint == "T1" ~ 9,
  alpha_DV_tr_m$Timepoint == "T2" ~ 14,
  alpha_DV_tr_m$Timepoint == "T3" ~ 19,
  TRUE ~ as.numeric(alpha_DV_tr_m$Timepoint)
)

alpha_DV_tr_m$Time_month <- case_when(
  alpha_DV_tr_m$Timepoint == "T0" ~ "March",
  alpha_DV_tr_m$Timepoint == "T1" ~ "June",
  alpha_DV_tr_m$Timepoint == "T2" ~ "July",
  alpha_DV_tr_m$Timepoint == "T3" ~ "August",
  TRUE ~ (alpha_DV_tr_m$Timepoint)
)

alpha_DV_tr_m$Time_month <- factor(alpha_DV_tr_m$Time_month, levels = c("March", "June", "July", "August"))

Fig2_C <- ggplot(alpha_DV_tr_m, aes(x = as.factor(Time_month), y = observed, fill = bulk_rhizo, color = bulk_rhizo)) +
  geom_boxplot(color = "black", alpha = 0.7, position = position_dodge(0.75)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.5, dodge.width = 0.75), alpha = 1, color = "black") +
  labs(x = "", y = "Number of vOTUs") +
  scale_fill_manual(name = "soil", values = compartment_colors) +
  scale_color_manual(name = "soil", values = compartment_colors) +
  theme_soil()


library(dplyr)
library(rstatix)

DV_richness_stat_test <- alpha_DV_tr_m %>%
  group_by(Time_month) %>%
  filter(n_distinct(bulk_rhizo) > 1) %>%
  wilcox_test(observed ~ bulk_rhizo) %>%
  add_significance("p") %>%
  add_xy_position(x = "Time_month", dodge = 0.8)

Fig2_C_stat <- Fig2_C +
  stat_pvalue_manual(
    DV_richness_stat_test,             # your stat_test dataframe
    label = "p.signif",    # shows *, **, etc.
    tip.length = 0.02,     # length of the whisker
    hide.ns = FALSE,        # hide non-significant comparisons
    position = position_dodge(width = 0.8),  # align with dodged boxes
    label.size = 8
  )

Fig2_C_stat


ggsave(paste0(workdir, "plots/alpha_div/Fig2_C_stat_DV_jitter_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_C_stat, 
       width = 7, height = 6, dpi = 600)

```


## Panel D

```{r}
library(readr)
metadata_AMF_all_11182025 <- read_csv("Library/CloudStorage/OneDrive-UniversityofCalifornia,Davis/Projects/Tomato_rhizosphere_Ani/Analysis/source_data/metadata_AMF_all_11182025.csv")

colnames(metadata_AMF_all_11182025)

metadata_AMF_all_11182025 <- metadata_AMF_all_11182025 |>
  dplyr::rename(Timepoint = Timeoint)

### Boxplots
metadata_AMF_all_11182025$bulk_rhizo_t <- paste0(
  metadata_AMF_all_11182025$bulk_rhizo, "_", metadata_AMF_all_11182025$Timepoint
)

metadata_AMF_all_11182025$bulk_rhizo_t <- factor(
  metadata_AMF_all_11182025$bulk_rhizo_t,
  levels = names(time_colors)
)


metadata_AMF_all_11182025$Time_month <- case_when(
  metadata_AMF_all_11182025$Timepoint == "T0" ~ "March",
  metadata_AMF_all_11182025$Timepoint == "T1" ~ "June",
  metadata_AMF_all_11182025$Timepoint == "T2" ~ "July",
  metadata_AMF_all_11182025$Timepoint == "T3" ~ "August",
  TRUE ~ (metadata_AMF_all_11182025$Timepoint)
)

metadata_AMF_all_11182025$Time_month <- factor(metadata_AMF_all_11182025$Time_month, levels = c("March", "June", "July", "August"))

library(dplyr)
library(rstatix)

DV_DNA_concentration_stat_test <- metadata_AMF_all_11182025 %>%
  group_by(Time_month) %>%
  filter(n_distinct(bulk_rhizo) > 1) %>%
  wilcox_test(DNA_concentration ~ bulk_rhizo) %>%
  add_significance("p") %>%
  add_xy_position(x = "Time_month", dodge = 0.8)


Fig2_D <- ggplot(metadata_AMF_all_11182025, aes(x = as.factor(Time_month), y = DNA_concentration, fill = bulk_rhizo, color = bulk_rhizo)) +
  geom_boxplot(color = "black", alpha = 0.7, position = position_dodge(0.75)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75), alpha = 1, color = "black") +
  scale_fill_manual(name = 'soil', values = compartment_colors) +
  scale_color_manual(name = 'soil', values = compartment_colors) +
  labs(x = "",y = "DNA concentration (ng/µL)") +theme_soil()


Fig2_D_sig <- Fig2_D +
  stat_pvalue_manual(
    DV_DNA_concentration_stat_test,             # your stat_test dataframe
    label = "p.signif",    # shows *, **, etc.
    tip.length = 0.02,     # length of the whisker
    hide.ns = FALSE,        # hide non-significant comparisons
    position = position_dodge(width = 0.8),  # align with dodged boxes
    label.size = 8
  )

Fig2_D_sig

ggsave(paste0(workdir, "plots/DNA_concentrations/DV_boxplots_DNA_concentration_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig2_D_sig, 
       width = 7, height = 6, dpi = 600)

```

## Combine Fig 2 ABCD

```{r}
library(patchwork)

top_right <- ((Fig2_B_T1_3 | Fig2_B_T2_3) /
                (Fig2_B_T3_3 | Fig2_B_T3_3)) +   # <-- wrap all 4 plots
  plot_layout(widths = c(1,1), heights = c(1,1), guides = "collect") & 
  theme(plot.margin = margin(2,2,2,2))

top_right
# Combine top row: Fig2_A on left, 4 small figs on right
top_row <- Fig2_A | top_right + plot_layout(widths = c(2, 1)) &
  theme(plot.margin = margin(2,2,2,2))

# Bottom row: C and D side by side
bottom_row <- Fig2_C_stat | Fig2_D_sig &
  theme(plot.margin = margin(2,2,2,2))

# Combine top and bottom rows
Figure_2_combined <- top_row / bottom_row +
  plot_layout(heights = c(1,1)) +  # same height for top & bottom
  plot_annotation(tag_levels = 'A')

Figure_2_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_2_combined_shorter_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_2_combined, 
      width = 14, height = 12, dpi = 600)

```

### Combined Fig 2 B

```{r}

# Combine with patchwork
Figure_2_B_combined <- 
                      (Fig2_B_T1_3 | Fig2_B_T2_3 | Fig2_B_T3_3) +
  plot_annotation(tag_levels = 'A')

Figure_2_B_combined

ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_2_B_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_2_B_combined, 
        width = 7, height = 6, dpi = 600)

```

# Figure 3

## Panel A


```{r}
## Split the data to different soil compartements

ps_DV_tr_bulk <- subset_samples(ps_DV_tr, bulk_rhizo == "bulk")
ps_DV_tr_bulk <- filter_taxa(ps_DV_tr_bulk, function(x) sum(x) > 0, TRUE)

pcoa_DV_tr_bulk <- ordinate(ps_DV_tr_bulk, "PCoA", distance = "bray", binary = F)

###### DNA viruses
ps_DV_tr_bulk@sam_data$Numeric_Timepoint <- case_when(
  ps_DV_tr_bulk@sam_data$Timepoint == "T0" ~ 0,
  ps_DV_tr_bulk@sam_data$Timepoint == "T1" ~ 9, # 6
  ps_DV_tr_bulk@sam_data$Timepoint == "T2" ~ 14, # 10
  ps_DV_tr_bulk@sam_data$Timepoint == "T3" ~ 19, # 17
  TRUE ~ as.numeric(ps_DV_tr_bulk@sam_data$Timepoint)
)

ps_DV_tr_bulk@sam_data$Time_month <- case_when(
  ps_DV_tr_bulk@sam_data$Timepoint == "T0" ~ "March",
  ps_DV_tr_bulk@sam_data$Timepoint == "T1" ~ "June",
  ps_DV_tr_bulk@sam_data$Timepoint == "T2" ~ "July",
  ps_DV_tr_bulk@sam_data$Timepoint == "T3" ~ "August",
  TRUE ~ (ps_DV_tr_bulk@sam_data$Timepoint)
)

# Visualization of PCoA results for each niche and coloring by region
pcoa_result_DV_tr_bulk <- as.data.frame(pcoa_DV_tr_bulk$vectors)

# Extract sample data for temporal information
temporal_data_DV_tr_bulk <- as.data.frame(sample_data(ps_DV_tr_bulk))

# Combine data for plotting
combined_data_DV_tr_bulk <- cbind(temporal_data_DV_tr_bulk, pcoa_result_DV_tr_bulk)

# Calculate centroids for each group and timepoint
centroid_data_DV_tr_bulk <- combined_data_DV_tr_bulk %>%
  group_by( Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

#########################################################
# Convert Timepoint to factor with the desired order
combined_data_DV_tr_bulk$Timepoint <- factor(combined_data_DV_tr_bulk$Timepoint, levels = c("T0", "T1", "T2", "T3"))

# Arrange data for plotting
arranged_data_DV_tr_bulk <- centroid_data_DV_tr %>%
  arrange( Timepoint)

# Figure 3 - panel A - with arrows
Fig3_A <- ggplot(combined_data_DV_tr_bulk, aes(x=Axis.1, y=Axis.2)) + 
  geom_point(aes(fill=soil_moisture_perc, shape=Time_month), color='black', size=6) +
  scale_fill_gradient(
    low = "#D0B17F",   # dry soil
    high = "#2171b5"   # wet soil
  ) +
  scale_shape_manual(
    name = "Time_month",
    breaks = c("March", "June", "July", "August"),
    values = c(21,22,23,24)
  ) +
    geom_text_repel(data = centroid_data_DV_tr_bulk, aes(x = Centroid_X, y = Centroid_Y, label = Timepoint),
                  size = 4, col = "black", fontface = 3) +
  theme_bw() +
  theme(
    text = element_text(size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_bulk$values$Eigenvalues[1] / sum(pcoa_DV_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_bulk$values$Eigenvalues[2] / sum(pcoa_DV_tr_bulk$values$Eigenvalues) * 100, 2), "%)")) +
  
  # Add arrows connecting centroids
  geom_segment(
    data = centroid_data_DV_tr_bulk,
    aes(
      x = lag(Centroid_X), y = lag(Centroid_Y),
      xend = Centroid_X, yend = Centroid_Y
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    color = "grey",
    na.rm = TRUE
  ) + theme_soil()


ggsave(paste0(workdir, "plots/pcoa/Fig3_A_moisture_pcoa_DV_tr_bulk_NO_LEGEND_arrows_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig3_A, 
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/Fig3_A_moisture_pcoa_DV_tr_bulk_NO_LEGEND_arrows_2_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig3_A, 
       width = 6, height = 5, dpi = 600)


## With legend

Fig3_A_legend <- ggplot(combined_data_DV_tr_bulk, aes(x=Axis.1, y=Axis.2)) + 
  geom_point(aes(fill=soil_moisture_perc, shape=Time_month), color='black', size=6) +
  scale_fill_gradient(
    low = "#D0B17F",   # dry soil
    high = "#2171b5"   # wet soil
  ) +
  scale_shape_manual(
    name = "Time_month",
    breaks = c("March", "June", "July", "August"),
    values = c(21,22,23,24)
  ) +
    geom_text_repel(data = centroid_data_DV_tr_bulk, aes(x = Centroid_X, y = Centroid_Y, label = Timepoint),
                  size = 4, col = "black", fontface = 3) +
  theme_bw()

ggsave(paste0(workdir, "plots/pcoa/Fig3_A_legend_moisture_pcoa_DV_tr_bulk_LEGEND_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig3_A_legend, 
       width = 6, height = 5, dpi = 600)

```

## Panel B

```{r}
## 6️⃣ Axis.1 vs Time
data_dv_bulk <- combined_data_DV_tr %>% filter(bulk_rhizo == "bulk") %>%
  filter(!is.na(Time_weeks) & !is.na(Axis.1))
cor_dv_time <- cor.test(data_dv_bulk$Axis.1, data_dv_bulk$Time_weeks)

Fig3_B <-   ggplot(data_dv_bulk, aes(x = Axis.1, y = Time_weeks, fill = soil_moisture_perc, shape = Time_month)) +
    geom_point(size = 6) +
   # geom_smooth(method = "lm", se = FALSE, color = "black") +
    geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  annotate(
    "text",
    x = Inf,
    y = -Inf,
    label = paste0(
      "r = ", round(cor_dv_time$estimate, 2),
      ", p = ", signif(cor_dv_time$p.value, 2)
    ),
    hjust = 1.1,
    vjust = -0.5,
    size = 6
  ) +
    moisture_gradient +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
    labs(x = "Bulk soil DNA viral PCoA Axis 1", y = "Time (weeks)", fill = "Soil moisture (%)") + 
    theme_soil()

Fig3_B


ggsave(paste0(workdir, "plots/axis_moisture/DV_Axis1_vs_Time_", Sys.Date(), ".pdf"),
       plot = Fig3_B, 
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/axis_moisture/DV_Axis1_vs_Time_2_", Sys.Date(), ".pdf"),
       plot = Fig3_B, 
       width = 6, height = 5, dpi = 600)


```

## Panel C

```{r}

cor_time_moisture <- cor.test(data_dv_bulk$soil_moisture_perc, data_dv_bulk$Time_weeks)

Fig3_C <- ggplot(data_dv_bulk, aes(x = Time_weeks, y = soil_moisture_perc, fill = soil_moisture_perc, shape = Time_month)) +
  geom_point(size = 6) +
    geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  annotate("text",
           x = min(data_dv_bulk$Time_weeks, na.rm = TRUE),
           y = max(data_dv_bulk$soil_moisture_perc, na.rm = TRUE),
           label = paste0("r = ", round(cor_time_moisture$estimate, 2),
                          ", p = ", signif(cor_time_moisture$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 4) +
    moisture_gradient_fill +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
  labs(x = "Time (weeks)", y = "Soil moisture (%)", color = "Sampling month") +
  theme_soil()


Fig3_C

ggsave(paste0(workdir, "plots/axis_moisture/plot_time_moisture_", Sys.Date(), ".pdf"),
       plot = Fig3_C, 
       width = 7, height = 6, dpi = 600)

```


## Panel D


```{r}

## Axis.1 vs Soil Moisture
data_dv_bulk <- combined_data_DV_tr %>% filter(bulk_rhizo == "bulk") %>%
  filter(!is.na(soil_moisture_perc) & !is.na(Axis.1))
cor_dv_soil <- cor.test(data_dv_bulk$soil_moisture_perc, data_dv_bulk$Axis.1)

Fig3_D <- ggplot(data_dv_bulk, aes(x = Axis.1, y = soil_moisture_perc, fill = soil_moisture_perc, shape = Time_month)) +
  geom_point(size = 6) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  annotate("text",x = Inf,y = -Inf,
    label = paste0(
      "r = ", round(cor_dv_soil$estimate, 2),
      ", p = ", signif(cor_dv_soil$p.value, 2)
    ), hjust = 1.1,vjust = -0.5,size = 6)+
    moisture_gradient_fill +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
  labs(x = "Bulk soil DNA viral PCoA Axis 1", y = "Soil moisture (%)", color = "Time (weeks)") +
  theme_soil()

Fig3_D

ggsave(paste0(workdir, "plots/axis_moisture/DV_Axis1_vs_SoilMoisture_", Sys.Date(), ".pdf"),
       plot = Fig3_D, 
       width = 7, height = 6, 
       dpi = 600)

ggsave(paste0(workdir, "plots/axis_moisture/DV_Axis1_vs_SoilMoisture_2_", Sys.Date(), ".pdf"),
       plot = Fig3_D, 
       width = 6, height = 5, 
       dpi = 600)
```

## Panel E

```{r}

## DNA viruses
Fig3_E <- 
  ggplot(alpha_DV_tr_m %>% filter( bulk_rhizo == "bulk"),
         aes(x = soil_moisture_perc, y = observed, fill = soil_moisture_perc, shape = Time_month)
  ) +
  geom_point(size = 6) +
  labs(
    title = "",
    x = "Soil Moisture (%)",
    y = "Number of vOTUs"
  ) +
  
  # Single regression line for all bulk samples
  geom_smooth(
    data = alpha_DV_tr_m %>% filter( bulk_rhizo == "bulk"),
    mapping = aes(x = soil_moisture_perc, y = observed), # explicitly set
    method = "lm",
    se = FALSE,
    color = "black",
    inherit.aes = FALSE # prevents re-using main plot aesthetics
  ) +
  
  # R² and p-value for all bulk samples
  stat_poly_eq(
    data = alpha_DV_tr_m %>% filter( bulk_rhizo == "bulk"),
    mapping = aes(x = soil_moisture_perc, y = observed,
                  label = paste(..rr.label.., ..p.value.label.., sep = "~~~")),
    formula = y ~ x,
    parse = TRUE,
    color = "black",
    inherit.aes = FALSE
  ) +
  
  # Manual color and shape scales
    moisture_gradient_fill +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
  theme_soil()


Fig3_E


ggsave(
  filename = paste0(workdir, "plots/richness_moisture/scatter_alpha_moisture_plot_DV_time_", Sys.Date(), ".pdf"),
  plot = Fig3_E,
  width = 7, height = 6
)

```

## Panel F

```{r}

Fig3_F <- ggplot(metadata_AMF_all_11182025 %>% filter(bulk_rhizo == "bulk"),
       aes(x = soil_moisture_erc,
           y = DNA_concentration,
           fill = soil_moisture_erc,
           shape = Time_month)) +
  geom_point(size = 6) +
    geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  scale_color_manual(values = bulk_month_colors) +
  scale_y_continuous(limits = c(0, 0.5)) +
  labs(
    x = "Soil Moisture (%)",
    y = "DNA Concentration (ng/µL)",
    color = "Timepoint"
  ) +
  # R² and p-value
  stat_poly_eq(data = metadata_AMF_all_11182025 %>% filter(bulk_rhizo == "bulk"),
    mapping = aes(
      x = soil_moisture_erc,
      y = DNA_concentration,
      label = paste(..rr.label.., ..p.value.label.., sep = "~~~")
    ),
    formula = y ~ x,
    parse = TRUE,
    color = "black",
    inherit.aes = FALSE
  ) +
    # Manual color and shape scales
    moisture_gradient_fill +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
  theme_soil()

ggsave(
  filename = paste0(workdir, "plots/DNA_concentrations/DNA_concentration_scatter_plot_bulk_soil", Sys.Date(), ".pdf"),
  plot = Fig3_F,
  width = 7, height = 6
)

```

## Panel G

```{r}
###############################################
### 1. Merge metadata
###############################################

ordered_sample_ids <- sample_data(ps_DV_tr_bulk)$SampleID

virus_data_ordered <- virus_data[ordered_sample_ids, ]

# Scale and Order Soil Chemistry Data
soil_chem_data <- meta(ps_DV_tr_bulk)
soil_chem_data <- soil_chem_data[ordered_sample_ids, ]

## Without soil moisture
#soil_chem_data_scaled <- soil_chem_data %>% 
#  as.data.frame() %>% 
#  select(-('WDRF_Buffer_pH')) %>%
#  mutate(across(`Soil_pH`:`CEC_Sum_of_Cations_me_100g`, ~scale(.))) %>%
#  select(`Soil_pH`:`CEC_Sum_of_Cations_me_100g`) 

## WITH soil moisture
soil_chem_data_scaled <- soil_chem_data %>% 
  as.data.frame() %>% 
  select(-('WDRF_Buffer_pH')) %>%
  mutate(across(`soil_moisture_perc`:`CEC_Sum_of_Cations_me_100g`, ~scale(.))) %>%
  select(`soil_moisture_perc`:`CEC_Sum_of_Cations_me_100g`) 

soil_chem_data_scaled$SampleID <- rownames(soil_chem_data_scaled)

soil_data_full <- merge(
  soil_chem_data_scaled,
  soil_chem_data[, 1:6],   
  by = "SampleID"
)

# Order samples by Timepoint & Plot (you can change if needed)
soil_data_full <- soil_data_full %>%
  arrange(Timepoint, Plot)

rownames(soil_data_full) <- soil_data_full$SampleID

soil_data_full$Time_month <- case_when(
  soil_data_full$Timepoint == "T0" ~ "March",
  soil_data_full$Timepoint == "T1" ~ "June",
  soil_data_full$Timepoint == "T2" ~ "July",
  soil_data_full$Timepoint == "T3" ~ "August",
  TRUE ~ (soil_data_full$Timepoint)
)

###############################################
### 2. Extract scaled chemistry matrix
###############################################

soil_matrix <- as.matrix(
  soil_data_full[ ,
                  which(names(soil_data_full) == "soil_moisture_perc"):
                    which(names(soil_data_full) == "CEC_Sum_of_Cations_me_100g")]
)


###############################################
### 3. Annotation dataframe (only Plot & Timepoint)
###############################################

annotation_df <- soil_data_full[, c("Plot", "Time_month")]


###############################################
### 4. Annotation colors
###############################################

annotation_colors <- list(
  Plot = plot_colors,
  Time_month = bulk_month_colors
)


###############################################
### 5. Neutral, non-conflicting heatmap palette
###############################################

heatmap_colors <- colorRampPalette(c("white", "#b9b2d8", "#7f78a8", "#3c3a4a"))(200)
# light lavender → medium purple → deep grey-purple


###############################################
### 6. Heatmap
###############################################


# Bulk Soil Chemistry (Sample-clustered)
Fig3_G <- pheatmap(
  soil_matrix,
  cluster_rows = TRUE,        # cluster samples
  cluster_cols = FALSE,       # keep variable order
  annotation_row = annotation_df,
  annotation_colors = annotation_colors,
  border_color = NA,
  color = heatmap_colors,
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 12,
) + theme_soil()


pheatmap(
  soil_matrix,
  cluster_rows = FALSE,        # cluster samples
  cluster_cols = TRUE,       # keep variable order
  annotation_row = annotation_df,
  annotation_colors = annotation_colors,
  border_color = NA,
  color = heatmap_colors,
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 12,
  main = "Bulk Soil Chemistry (Variable-clustered)"
)



pdf(paste0(workdir, "plots/heatmap/soil_chemistry_heatmap_", Sys.Date(), ".pdf"), width = 15, height = 10)

pheatmap(
  soil_matrix,
  cluster_rows = TRUE,        # cluster samples
  cluster_cols = FALSE,       # keep variable order
  annotation_row = annotation_df,
  annotation_colors = annotation_colors,
  border_color = NA,
  color = heatmap_colors,
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 24,
  main = "Bulk Soil Chemistry (Sample-clustered)"
)

dev.off()



pdf(paste0(workdir, "plots/heatmap/soil_chemistry_heatmap_varaible_clustered_", Sys.Date(), ".pdf"), width = 10, height = 8)

pheatmap(
  soil_matrix,
  cluster_rows = FALSE,        # cluster samples
  cluster_cols = TRUE,       # keep variable order
  annotation_row = annotation_df,
  annotation_colors = annotation_colors,
  border_color = NA,
  color = heatmap_colors,
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize = 12,
  main = "Bulk Soil Chemistry (Variable-clustered)"
)

dev.off()


```

#### Soil chemistry correlation and PCA

```{r}
# Load required packages
library(tidyverse)
library(ggfortify)  # for autoplot PCA
library(vegan)      # for scaling if needed

soil_chem_data_scaled_NO_moisture <- soil_chem_data_scaled %>%
  select(-soil_moisture_perc)   # replace with the exact column name if different

# Run PCA
pca_chem <- prcomp(soil_chem_data_scaled_NO_moisture, center = TRUE, scale. = TRUE)

# Summary of PCA
summary(pca_chem)  # variance explained by PC1, PC2, etc.

# Extract scores for samples
scores <- as.data.frame(pca_chem$x)

# Add metadata (time and soil moisture)
# If there is a column for bulk/rhizo, filter bulk only
temporal_df_bulk <- temporal_data_DV_tr[temporal_data_DV_tr$bulk_rhizo == "bulk", ]

# Now combine with PCA scores
pca_soil_chem_data <- cbind(temporal_df_bulk, scores)

pca_soil_chem_data

pca_soil_chem_data$Time_month <- case_when(
  pca_soil_chem_data$Timepoint == "T0" ~ "March",
  pca_soil_chem_data$Timepoint == "T1" ~ "June",
  pca_soil_chem_data$Timepoint == "T2" ~ "July",
  pca_soil_chem_data$Timepoint == "T3" ~ "August",
  TRUE ~ (pca_soil_chem_data$Timepoint)
)

pca_soil_chem_data$Time_weeks <- case_when(
  pca_soil_chem_data$Timepoint == "T0" ~ "0",
  pca_soil_chem_data$Timepoint == "T1" ~ "9",
  pca_soil_chem_data$Timepoint == "T2" ~ "14",
  pca_soil_chem_data$Timepoint == "T3" ~ "19",
  TRUE ~ (pca_soil_chem_data$Timepoint)
)

pca_soil_chem_data$Time_weeks <- as.numeric(pca_soil_chem_data$Time_weeks)
# Correlate PC1 with time and soil moisture
cor.test(pca_soil_chem_data$PC1, pca_soil_chem_data$Time_weeks)        # correlation with linear time
cor.test(pca_soil_chem_data$PC1, pca_soil_chem_data$soil_moisture_perc)    # correlation with soil moisture

# Scatter plot: PC1 vs Soil Moisture
cor_Fig3_G_2 <- cor.test(pca_soil_chem_data$soil_moisture_perc, pca_soil_chem_data$PC1)

Fig3_G_2 <- 
   ggplot(pca_soil_chem_data, aes(x = soil_moisture_perc, y = PC1, fill = soil_moisture_perc, shape = Time_month)) +
  geom_point(size = 6) +
    geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +
  annotate("text",
           x = min(pca_soil_chem_data$soil_moisture_perc, na.rm = TRUE),
           y = max(pca_soil_chem_data$PC1, na.rm = TRUE),
           label = paste0("r = ", round(cor_Fig3_G_2$estimate, 2),
                          ", p = ", signif(cor_Fig3_G_2$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 4) +
    moisture_gradient +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
  labs(x = "Soil moisture (%)", y = "PC1 (Soil Chemistry)", color = "Soil moisture (%)") + 
    theme_soil()



Fig3_G_2

ggsave(paste0(workdir, "plots/axis_moisture/PCA_BulkChem_SoilMoisture_", Sys.Date(), ".pdf"),
       plot = Fig3_G_2, 
       width = 7, height = 6, 
       dpi = 600)

####
cor_Fig3_G_3 <- cor.test(pca_soil_chem_data$Time_weeks, pca_soil_chem_data$PC1)

Fig3_G_3 <- ggplot(pca_soil_chem_data, aes(x = Time_weeks, y = PC1, color = Time_month)) +
  geom_point(size = 6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  annotate("text",
           x = min(pca_soil_chem_data$Time_weeks, na.rm = TRUE),
           y = max(pca_soil_chem_data$PC1, na.rm = TRUE),
           label = paste0("r = ", round(cor_Fig3_G_3$estimate, 2),
                          ", p = ", signif(cor_Fig3_G_3$p.value, 2)),
           hjust = 0, vjust = 1.5, size = 4) +
  labs(x = "Time (weeks)", y = "PC1 (Soil Chemistry)") +
  scale_color_manual(values = c("March" = "#C9A46C", "June" = "#BA7F3A",
                                "July" = "#8F5A28", "August" = "#5E4224")) +
  theme_soil()

Fig3_G_3

ggsave(paste0(workdir, "plots/axis_moisture/PCA_BulkChem_Time_", Sys.Date(), ".pdf"),
       plot = Fig3_G_3, 
       width = 7, height = 6, 
       dpi = 600)

### PCA

library(ggplot2)
library(ggfortify)  # for autoplot if needed

# Make Time_month a factor
pca_soil_chem_data$Time_month <- factor(pca_soil_chem_data$Time_month, levels = c("March", "June", "July", "August"))

# Basic PCA biplot with convex hulls by Time_month
Fig3_G_4 <- ggplot(pca_soil_chem_data, aes(x = PC1, y = PC2, color = Time_month, color = Time_month)) +
  geom_point(size = 6, alpha = 1) +
  
  # Convex hulls for each time point
  geom_polygon(data = pca_soil_chem_data %>% group_by(Time_month) %>% slice(chull(PC1, PC2)),
               aes(x = PC1, y = PC2, fill = Time_month),
               alpha = 0.1, inherit.aes = TRUE) +
  xlab(paste0("PC1 (", round(summary(pca_chem)$importance[2,1]*100, 1), "%)")) +
  ylab(paste0("PC2 (", round(summary(pca_chem)$importance[2,2]*100, 1), "%)")) +
  scale_color_manual(values = c("March" = "#C9A46C", "June" = "#BA7F3A",
                                "July" = "#8F5A28", "August" = "#5E4224")) +
  
    scale_fill_manual(values = c("March" = "#C9A46C", "June" = "#BA7F3A",
                                "July" = "#8F5A28", "August" = "#5E4224")) +
  theme_soil()


Fig3_G_4

ggsave(paste0(workdir, "plots/axis_moisture/PCA_BulkChemistry_", Sys.Date(), ".pdf"),
       plot = Fig3_G_4, 
       width = 7, height = 6, 
       dpi = 600)


### PCoA Axis 1 and PCA axis 1

############
# Check presence
all(c("SampleID") %in% colnames(pca_soil_chem_data))
all(c("SampleID") %in% colnames(data_dv_bulk))

# Check if they are in the same order
identical(
  pca_soil_chem_data$SampleID,
  data_dv_bulk$SampleID
)

# Check for missing
setdiff(pca_soil_chem_data$SampleID, data_dv_bulk$SampleID)
setdiff(data_dv_bulk$SampleID, pca_soil_chem_data$SampleID)

# Join

pca_soil_chem_DV_data <- pca_soil_chem_data %>%
  left_join(
    data_dv_bulk %>% 
      select(SampleID, Axis.1, Axis.2),
    by = "SampleID"
  )


cor_Fig3_G_5 <- cor.test(pca_soil_chem_DV_data$Axis.1, pca_soil_chem_DV_data$PC1)

Fig3_G_5 <- ggplot(pca_soil_chem_DV_data, aes(x = Axis.1, y = PC1, fill = soil_moisture_perc, shape = Time_month)) +
  geom_point(size = 6) +
    geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "black") +annotate(
    "text",
    x = Inf,
    y = -Inf,
    label = paste0(
      "r = ", round(cor_Fig3_G_5$estimate, 2),
      ", p = ", signif(cor_Fig3_G_5$p.value, 2)
    ),
    hjust = 1.1,
    vjust = -0.5,
    size = 6
  )+
    moisture_gradient_fill +
    scale_shape_manual(
      name = "Time_month",
      breaks = c("March", "June", "July", "August"),
      values = c(21,22,23,24)
    ) +
  labs(x = "Bulk soil DNA viral PCoA Axis 1", y = "Bulk soil chemistry PCA Axis 1") +
  theme_soil()

Fig3_G_5

ggsave(paste0(workdir, "plots/axis_moisture/cor_DV_PcoA_soilchem_PCA_", Sys.Date(), ".pdf"),
       plot = Fig3_G_5, 
       width = 7, height = 6, 
       dpi = 600)

ggsave(paste0(workdir, "plots/axis_moisture/cor_DV_PcoA_soilchem_PCA_2_", Sys.Date(), ".pdf"),
       plot = Fig3_G_5, 
       width = 6, height = 5, 
       dpi = 600)

```


## Combine Fig 3 ABCDEF


```{r}
# Fig3_A # Fig3_B # Fig3_D # Fig3_G_5

library(patchwork)

Figure_3_combined <-
  (Fig3_A | Fig3_D) /
  (Fig3_B | Fig3_G_5) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.margin = margin(5, 5, 5, 5)
  )

Figure_3_combined


# Combine with patchwork
Figure_3_combined <- (Fig3_A | Fig3_B) / (Fig3_D | Fig3_G_5 )+
  plot_annotation(tag_levels = 'A')

Figure_3_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_3_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_3_combined, 
       width = 12, height = 10, dpi = 600)

# Add Panel G manually

```


# Figure 4


## Panel A

```{r}

ps_DV_tr_rhizo <- subset_samples(ps_DV_tr, bulk_rhizo == "rhizo")
ps_DV_tr_rhizo <- filter_taxa(ps_DV_tr_rhizo, function(x) sum(x) > 0, TRUE)


pcoa_DV_tr_rhizo <- ordinate(ps_DV_tr_rhizo, "PCoA", distance = "bray", binary = F)

###### DNA viruses
ps_DV_tr_rhizo@sam_data$Numeric_Timepoint <- case_when(
  #ps_DV_tr_rhizo@sam_data$Timepoint == "T0" ~ 0,
  ps_DV_tr_rhizo@sam_data$Timepoint == "T1" ~ 9, # 6
  ps_DV_tr_rhizo@sam_data$Timepoint == "T2" ~ 14, # 10
  ps_DV_tr_rhizo@sam_data$Timepoint == "T3" ~ 19, # 17
  TRUE ~ as.numeric(ps_DV_tr_rhizo@sam_data$Timepoint)
)

ps_DV_tr_rhizo@sam_data$Time_month <- case_when(
  #ps_DV_tr_rhizo@sam_data$Timepoint == "T0" ~ "March",
  ps_DV_tr_rhizo@sam_data$Timepoint == "T1" ~ "June",
  ps_DV_tr_rhizo@sam_data$Timepoint == "T2" ~ "July",
  ps_DV_tr_rhizo@sam_data$Timepoint == "T3" ~ "August",
  TRUE ~ (ps_DV_tr_rhizo@sam_data$Timepoint)
)

# Visualization of PCoA results for each niche and coloring by region
pcoa_result_DV_tr_rhizo <- as.data.frame(pcoa_DV_tr_rhizo$vectors)

# Extract sample data for temporal information
temporal_data_DV_tr_rhizo <- as.data.frame(sample_data(ps_DV_tr_rhizo))

# Combine data for plotting
combined_data_DV_tr_rhizo <- cbind(temporal_data_DV_tr_rhizo, pcoa_result_DV_tr_rhizo)

# Calculate centroids for each group and timepoint
centroid_data_DV_tr_rhizo <- combined_data_DV_tr_rhizo %>%
  group_by( Timepoint) %>%
  summarize(Centroid_X = mean(Axis.1), Centroid_Y = mean(Axis.2)) %>%
  ungroup()

#########################################################
# Convert Timepoint to factor with the desired order
combined_data_DV_tr_rhizo$Timepoint <- factor(combined_data_DV_tr_rhizo$Timepoint, levels = c("T0", "T1", "T2", "T3"))


plot_colors <- c(
  "1_4" = "#6FB7B4",  # pale lavender
  "5_5" = "#7A2E1D",  # muted teal
  "7_8" = "#8F7A3E"   # cool slate gray
)

amf_colors <- c(
  "Control"   = "#8C8C8C",
  "Treatment" = "#2F3E5C"   # deep navy
)

Fig4_A <- ggplot(combined_data_DV_tr_rhizo, aes(x = Axis.1, y = Axis.2, shape = Timepoint)) +
  geom_point(aes(fill = Plot), size = 6, color = "black") +
  geom_convexhull(aes(color = Plot,fill = Plot, group = Plot), alpha = 0.1) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_rhizo$values$Eigenvalues[1] / sum(pcoa_DV_tr_rhizo$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_rhizo$values$Eigenvalues[2] / sum(pcoa_DV_tr_rhizo$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()


ggsave(paste0(workdir, "plots/pcoa/pcoa_DV_rhizo_plots_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_A,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/Fig4_A_pcoa_DV_rhizo_plots_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_A,
       width = 6, height = 5, dpi = 600)

```


## Panel B


```{r}

# Suppose your CAP object is:
cap_RV_tr_rhizo <- ordinate(ps_RV_tr, method = "CAP", distance = "bray", formula = ~ Plot)
# Extract eigenvalues
eig_vals_cap_RV_tr_rhizo <- cap_RV_tr_rhizo$CCA$eig

# Percent variance explained by each constrained axis
cap_var_RV_tr_rhizo <- eig_vals_cap_RV_tr_rhizo / sum(eig_vals_cap_RV_tr_rhizo) * 100
cap_var_RV_tr_rhizo
# CAP1: cap_var[1], CAP2: cap_var[2]

# Then use in ggplot axis labels:
# 
# Fig4_B <- plot_ordination(ps_RV_tr, cap_RV_tr_rhizo, type = "samples", color = "Plot",  shape = "Timepoint") +
#   geom_point(size = 6, alpha = 1) +
#   geom_convexhull(aes(fill = Plot, group = Plot),alpha = 0.1 ) +
#   scale_color_manual(values = plot_colors) +
#   scale_fill_manual(values = plot_colors) +
#   scale_shape_manual(
#     name = "Timepoint",
#     breaks = c("T1", "T2", "T3"),
#     values = c(22,23,24)) +
#   xlab(paste0("CAP1 (", round(cap_var_RV_tr_rhizo[1], 1), "%)")) +
#   ylab(paste0("CAP2 (", round(cap_var_RV_tr_rhizo[2], 1), "%)")) +
#   theme_soil()


Fig4_B <- plot_ordination(ps_RV_tr, cap_RV_tr_rhizo, type = "samples", color = "Plot",  shape = "Timepoint") +
  geom_point(aes(fill = Plot), size = 6, alpha = 1, color = "black") +  # border black
  geom_convexhull(aes(fill = Plot, group = Plot), alpha = 0.1) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_shape_manual(
    name = "Timepoint",
    breaks = c("T1", "T2", "T3"),
    values = c(22,23,24)) +
  xlab(paste0("CAP1 (", round(cap_var_RV_tr_rhizo[1], 1), "%)")) +
  ylab(paste0("CAP2 (", round(cap_var_RV_tr_rhizo[2], 1), "%)")) +
  theme_soil()


Fig4_B

ggsave(paste0(workdir, "plots/CAP/CAP_plot_RV_space_rhizo_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_B, 
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/CAP/Fig4_B_CAP_plot_RV_space_rhizo_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_B, 
       width = 6, height = 5, dpi = 600)

## Test the significance?

# Perform permutation test (anova)
anova(cap_RV_tr_rhizo)                  # Overall model significance
anova(cap_RV_tr_rhizo, by = "terms")    # Tests each predictor
anova(cap_RV_tr_rhizo, by = "axis")     # Tests each constrained axis

```


## Panel C

### RNAseq and deseq2

```{r}
###########################################################################
# DESeq2 Differential Expression + Functional Annotation
###########################################################################

# ------------------------------
# 0. Setup
# ------------------------------
set.seed(29092020)
dir.create(file.path(workdir, "plots/deseq2"), recursive = TRUE, showWarnings = FALSE)

# ------------------------------
# 1. Load Libraries
# ------------------------------
library(DESeq2)
library(dplyr)
library(readr)
library(tibble)
library(stringr)
library(apeglm)
library(pheatmap)
library(ggplot2)
library(tidyr)


# ------------------------------
# 2. Load Data
# ------------------------------

## 2a. Counts table
counts_file <- file.path(workdir, "RNA/merged_numreads.tsv")
merged_numreads <- read_delim(counts_file, delim = "\t", escape_double = FALSE, trim_ws = TRUE)
counts <- as.data.frame(merged_numreads)
rownames(counts) <- counts$Name
counts$Name <- NULL
counts <- round(counts)

## 2b. Metadata
metadata_file <- file.path(workdir, "source_data/metadata_AMF_rhizo.csv")
metadata_AMF_rhizo <- read_csv(metadata_file)
metadata <- metadata_AMF_rhizo %>% filter(SampleID %in% colnames(counts))

## 2c. EggNOG annotations
eggnog_file <- file.path(workdir, "RNA/eggnog_mapper.emapper.annotations")
eggnog_mapper_emapper <- read_tsv(
  eggnog_file,
  comment = "#",
  col_names = c(
    "query", "seed_ortholog", "evalue", "score", "eggNOG_OGs",
    "max_annot_lvl", "COG_category", "Description", "Preferred_name",
    "GOs", "EC", "KEGG_ko", "KEGG_Pathway", "KEGG_Module",
    "KEGG_Reaction", "KEGG_rclass", "BRITE", "KEGG_TC", "CAZy",
    "BiGG_Reaction", "PFAMs"
  )
)

# Remove isoform suffix (.p1, .p2)
library(dplyr)
library(stringr)

emapper.annotations <- eggnog_mapper_emapper %>% mutate(query = str_remove(query, "\\.p[0-9]+$"))

# Keep best annotation per transcript (lowest e-value)
emapper_best <- emapper.annotations %>%
  group_by(query) %>%
  slice_min(order_by = evalue, n = 1) %>%
  ungroup()

# ------------------------------
# 3. Build DESeq2 Dataset
# ------------------------------
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData = metadata,
  design = ~ Site + Timepoint + Treatment
)

# ------------------------------
# 4. Run DESeq2
# ------------------------------
dds <- DESeq(dds)

# Get DE results for Treatment vs Control
res <- results(dds, contrast = c("Treatment", "Treatment", "Control"))

summary(res, alpha = 0.05)

# Shrink log2 fold-changes for visualization
# https://introtogenomics.readthedocs.io/en/latest/2021.11.11.DeseqTutorial.html
resLFC <- lfcShrink(dds, coef = "Treatment_Treatment_vs_Control", type = "apeglm")

# Order results by adjusted p-value
resOrdered <- resLFC[order(resLFC$padj), ]
resOrdered.df <- as.data.frame(resOrdered) %>% tibble::rownames_to_column("transcript_id")

# Export DE results
write.csv(resOrdered.df, file.path(workdir, "results/DE_transcripts_AMF_vs_Control.csv"), row.names = FALSE)


# ------------------------------
# 4b. Variance-stabilizing transformation (for PCA, clustering)
# ------------------------------
vsd <- vst(dds, blind = TRUE)

# PCA plot
pca_data <- plotPCA(vsd, intgroup = c("Timepoint", "Treatment", "Site"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, color = Site, shape = Timepoint)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Variance-Stabilized Transcript Counts",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance")
  )

ggplot(pca_data, aes(PC1, PC2, color = Treatment, shape = Site)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Variance-Stabilized Transcript Counts",
    x = paste0("PC1: ", percentVar[1], "% variance"),
    y = paste0("PC2: ", percentVar[2], "% variance")
  )

ggsave(file.path(workdir, "plots/deseq2/PCA_vsd.pdf"), height = 5, width = 6)


# ------------------------------
# 5. Merge DE results with functional annotation
# ------------------------------
DE_annotated <- resOrdered.df %>%
  left_join(emapper_best, by = c("transcript_id" = "query"))

# Filter DEGs: p-adj < 0.05 & |log2FC| > 1
DE_mod_df <- DE_annotated %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
  mutate(direction = ifelse(log2FoldChange > 0, "Up", "Down"))

# Summary of up/down DEGs
DE_mod_df %>% dplyr::count(direction)

write.csv(DE_mod_df, paste0(workdir, "results/DE_filtered_results_",Sys.Date(), ".csv"))

# ------------------------------
# 6. Visualization
# ------------------------------

## 6a. Histogram of log2 fold changes
ggplot(resOrdered.df, aes(x = log2FoldChange)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of log2 Fold Changes", x = "log2 Fold Change", y = "Number of transcripts")

## 6b. Volcano plot
res_df_clean <- resOrdered.df %>% filter(!is.na(log2FoldChange) & !is.na(padj))
res_df_clean <- res_df_clean %>%
  mutate(sig = case_when(
    padj < 0.05 & abs(log2FoldChange) > 2 ~ "Strong",
    padj < 0.05 & abs(log2FoldChange) > 1 ~ "Moderate",
    padj < 0.05 ~ "Weak",
    TRUE ~ "Not Significant"
  ))

ggplot(res_df_clean, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(values = c("Strong"="red","Moderate"="orange","Weak"="yellow","Not Significant"="gray")) +
  theme_minimal() +
  geom_vline(xintercept = c(-1,1), linetype="dashed", color="blue") +
  geom_vline(xintercept = c(-2,2), linetype="dashed", color="red") +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color="black") +
  labs(title = "Volcano Plot", x = "log2 Fold Change", y = "-log10(padj)")

## 6c. Heatmap of top 50 DEGs
norm_counts <- counts(dds, normalized = TRUE)
top_transcripts <- DE_mod_df %>% arrange(desc(abs(log2FoldChange))) %>% slice_head(n = 50) %>% pull(transcript_id)
heatmat_log <- log2(norm_counts[top_transcripts, ] + 1)

# Build column annotation
metadata_AMF_rhizo <-  as.data.frame(metadata_AMF_rhizo)
rownames(metadata_AMF_rhizo) <- metadata_AMF_rhizo$SampleID
annotation_col_df <- metadata_AMF_rhizo[colnames(heatmat_log), c("Timepoint","Treatment")]

# Build transcript labels
transcript_annotation <- DE_mod_df %>% filter(transcript_id %in% rownames(heatmat_log)) %>%
  mutate(label = paste0(COG_category," (",sig,")")) %>%
  select(transcript_id,label) %>% tibble::column_to_rownames("transcript_id")
rownames(heatmat_log) <- transcript_annotation[rownames(heatmat_log), "label"]

pheatmap(
  heatmat_log,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  scale = "row",
  color = colorRampPalette(c("blue","white","red"))(50),
  main = "Top 50 DE Transcripts - Normalized Counts",
  annotation_col = annotation_col_df,
  filename = file.path(workdir, paste0("plots/deseq2/Top50_DE_Transcripts_", Sys.Date(), ".pdf")),
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 6,
  fontsize_col = 8,
  border_color = NA
)

## 6d. Functional category barplots (COG)
cog_counts <- DE_mod_df %>%
  filter(!is.na(COG_category)) %>%
  group_by(COG_category, direction) %>%
  summarise(n = n(), .groups = "drop")

ggplot(cog_counts, aes(x = COG_category, y = n, fill = direction)) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_manual(values=c("Up"="red","Down"="blue")) +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  labs(title="DE transcripts by COG category", x="COG category", y="Number of transcripts")

# ------------------------------
# 8. PERMANOVA on normalized counts
# ------------------------------
library(vegan)
set.seed(123)

vst_counts <- assay(vst(dds, blind = FALSE))
comm_matrix_vst <- t(vst_counts)

metadata <- metadata_AMF_rhizo[match(rownames(comm_matrix_vst), metadata_AMF_rhizo$SampleID), ]

dist_mat_vst <- vegdist(comm_matrix_vst, method = "euclidean")  # Use Euclidean with VST

adonis2(dist_mat_vst ~ Timepoint*Treatment*Site, data = metadata, permutations = 999, by = "terms")

# ------------------------------
# 7. Functional exploration (metatranscriptome-safe)
# ------------------------------

# A. COG category summaries (already partially done)
cog_summary <- DE_mod_df %>%
  filter(!is.na(COG_category)) %>%
  group_by(COG_category, direction) %>%
  summarise(n = n(), .groups = "drop")

write.csv(cog_summary, file.path(workdir, "results/COG_summary.csv"), row.names = FALSE)


# B. Broad functional category heatmap
COG_mat <- DE_mod_df %>%
  filter(!is.na(COG_category)) %>%
  dplyr::count(COG_category, direction) %>%
  tidyr::pivot_wider(names_from = direction, values_from = n, values_fill = 0)

write.csv(COG_mat, file.path(workdir, "results/COG_heatmap_data.csv"), row.names = FALSE)


# C. Transcript-level clustering using VST data
vsd_mat <- assay(vsd)
top_var <- head(order(rowVars(vsd_mat), decreasing = TRUE), 100)

pheatmap(
  vsd_mat[top_var, ],
  scale = "row",
  annotation_col = metadata_AMF_rhizo[, c("Timepoint", "Treatment", "Site")],
  main = "Top 100 Most Variable Transcripts (VST)",
  filename = file.path(workdir, "plots/deseq2/Top100_vsd_heatmap.pdf"),
  border_color = NA
)

####### Top 20
library(dplyr)
library(tibble)

# Convert DESeq2 results to a tibble
resLFC.df <- as.data.frame(resLFC) %>% tibble::rownames_to_column("transcript_id")

# Merge with annotation
resLFC_annot <- resLFC.df %>%
  left_join(emapper_best, by = c("transcript_id" = "query"))

# Filter significant DEGs (adjusted p < 0.05)
DEGs <- resLFC_annot %>% filter(padj < 0.05)

# Top 20 Up-regulated
top20_up <- DEGs %>%
  filter(log2FoldChange > 0) %>%
  arrange(desc(log2FoldChange)) %>%
  slice_head(n = 20) %>%
  select(transcript_id, log2FoldChange, padj, COG_category, Description, KEGG_ko)

# Top 20 Down-regulated
top20_down <- DEGs %>%
  filter(log2FoldChange < 0) %>%
  arrange(log2FoldChange) %>%  # most negative first
  slice_head(n = 20) %>%
  select(transcript_id, log2FoldChange, padj, COG_category, Description, KEGG_ko)

# Quick summary: annotated vs unannotated
top20_up %>% mutate(annotated = ifelse(is.na(COG_category) & is.na(Description), "No", "Yes")) %>%
  count(annotated)

top20_down %>% mutate(annotated = ifelse(is.na(COG_category) & is.na(Description), "No", "Yes")) %>%
  count(annotated)

# Export if desired
write.csv(top20_up, file.path(workdir, "Top20_Up_DEGs.csv"), row.names = FALSE)
write.csv(top20_down, file.path(workdir, "Top20_Down_DEGs.csv"), row.names = FALSE)

```


```{r}

# Use vsd as is your variance-stabilized DESeq2 object
vsd <- vst(dds, blind = TRUE)

# PCA data for plotting
pca_data <- plotPCA(vsd, intgroup = c("Timepoint", "Treatment", "Site"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

# Make SamplingPlot a factor with desired order
pca_data$SamplingPlot <- factor(pca_data$Site, levels = c("1_4", "5_5", "7_8")) # adjust

Fig4_C <- ggplot(pca_data, aes(PC1, PC2, color = SamplingPlot)) +
  geom_point(size = 6, alpha = 1) +
  #facet_wrap(~Timepoint) +
  geom_convexhull(aes(fill = SamplingPlot), alpha = 0.1) +
  scale_color_manual(values = plot_colors) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_soil()

Fig4_C


ggsave(paste0(workdir, "plots/metaT/PCA_metaT_time_samplingplot_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C, 
       width = 7, height = 6, dpi = 600)



```


### Timepoints separate:

```{r}

dds_T1 <- dds[, dds$Timepoint == "T1"]
dds_T2 <- dds[, dds$Timepoint == "T2"]
dds_T3 <- dds[, dds$Timepoint == "T3"]

vsd_T1 <- vst(dds_T1, blind = TRUE)
vsd_T2 <- vst(dds_T2, blind = TRUE)
vsd_T3 <- vst(dds_T3, blind = TRUE)

pca_T1 <- plotPCA(vsd_T1,
                  intgroup = c("Site", "Treatment"),
                  returnData = TRUE)

pca_T2 <- plotPCA(vsd_T2,
                  intgroup = c("Site", "Treatment"),
                  returnData = TRUE)

pca_T3 <- plotPCA(vsd_T3,
                  intgroup = c("Site", "Treatment"),
                  returnData = TRUE)

var_T1 <- round(100 * attr(pca_T1, "percentVar"))
var_T2 <- round(100 * attr(pca_T2, "percentVar"))
var_T3 <- round(100 * attr(pca_T3, "percentVar"))

## Plots
# Timepoint 1 = circle = 21
# Timepoint 2 = triangle = 24
# Timepoint 3 = square = 22

  scale_shape_manual(
    name = "Time_month",
    breaks = c("March", "June", "July", "August"),
    values = c(21,22,23,24)
  ) 
  
  
Fig4_C_T1 <- ggplot(pca_T1, aes(PC1, PC2, color = Site, fill = Site, shape = Timepoint)) +
  geom_point(aes(fill = Site), size = 6, color = "black") +
  geom_convexhull(aes(fill = Site, group = Site),alpha = 0.1 ) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  xlab(paste0("PC1: ", var_T1[1], "%")) +
  ylab(paste0("PC2: ", var_T1[2], "%")) +
  theme_soil()+
    scale_shape_manual(
    name = "Timepoint",
    breaks = c("T1", "T2", "T3"),
    values = c(22,23,24)
  ) +   theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

Fig4_C_T2 <- ggplot(pca_T2, aes(PC1, PC2, color = Site, fill = Site, shape = Timepoint)) +
  geom_point(aes(fill = Site), size = 6, color = "black") +
  geom_convexhull(aes(fill = Site, group = Site),alpha = 0.1 ) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  xlab(paste0("PC1: ", var_T2[1], "%")) +
  ylab(paste0("PC2: ", var_T2[2], "%")) +
  theme_soil()+
    scale_shape_manual(
    name = "Timepoint",
    breaks = c("T1", "T2", "T3"),
    values = c(22,23,24)
  ) +   theme(
    axis.text = element_text(size = 10), #12
    axis.title = element_text(size = 12) #14
  )

Fig4_C_T3 <- ggplot(pca_T3, aes(PC1, PC2,color = Site, fill = Site, shape = Timepoint)) +
  geom_point(aes(fill = Site), size = 6, color = "black") +
  geom_convexhull(aes(fill = Site, group = Site),alpha = 0.1 ) +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_color_manual(values = plot_colors) +
  xlab(paste0("PC1: ", var_T3[1], "%")) +
  ylab(paste0("PC2: ", var_T3[2], "%")) +
  theme_soil()+
    scale_shape_manual(
    name = "Timepoint",
    breaks = c("T1", "T2", "T3"),
    values = c(22,23,24)
  ) +   theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )


ggsave(paste0(workdir, "plots/metaT/Fig4_C_T1_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_T1, 
       width = 5, height = 4, dpi = 600)


ggsave(paste0(workdir, "plots/metaT/Fig4_C_T2_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_T2, 
       width = 5, height = 4, dpi = 600)


ggsave(paste0(workdir, "plots/metaT/Fig4_C_T3_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_T3, 
       width = 5, height = 4, dpi = 600)

######

ggsave(paste0(workdir, "plots/metaT/Fig4_C_T1_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_T1, 
       width = 2, height = 5, dpi = 600)


ggsave(paste0(workdir, "plots/metaT/Fig4_C_T2_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_T2, 
       width = 2, height =5, dpi = 600)


ggsave(paste0(workdir, "plots/metaT/Fig4_C_T3_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_T3, 
       width = 2, height = 5, dpi = 600)

```

Combined

```{r}

# Combine with patchwork
Figure_4_C_combined <- (Fig4_C_T1 | Fig4_C_T2 | Fig4_C_T3) + plot_annotation(tag_levels = 'A')


Figure_4_C_combined <- 
  (Fig4_C_T1 | Fig4_C_T2 | Fig4_C_T3) &
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )


Figure_4_C_combined


ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_4_C_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_4_C_combined, 
      width = 6, height = 5, dpi = 600)


```


```{r}

# Find the combined range first
xlim <- range(c(Fig4_C_T1$data$PC1, Fig4_C_T2$data$PC1, Fig4_C_T3$data$PC1))
ylim <- range(c(Fig4_C_T1$data$PC2, Fig4_C_T2$data$PC2, Fig4_C_T3$data$PC2))

# Update each plot to use same limits
Fig4_C_T1 <- Fig4_C_T1 + coord_fixed(xlim = xlim, ylim = ylim)
Fig4_C_T2 <- Fig4_C_T2 + coord_fixed(xlim = xlim, ylim = ylim)
Fig4_C_T3 <- Fig4_C_T3 + coord_fixed(xlim = xlim, ylim = ylim)


Figure_4_C_combined <- 
  (Fig4_C_T1 | Fig4_C_T2 | Fig4_C_T3) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_4_C_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_4_C_combined, 
      width = 6, height = 10, dpi = 600)


```


### Locations separate


```{r}

dds_1_4 <- dds[, dds$Site == "1_4"]
dds_5_5 <- dds[, dds$Site == "5_5"]
dds_7_8 <- dds[, dds$Site == "7_8"]

vsd_1_4 <- vst(dds_1_4, blind = TRUE)
vsd_5_5 <- vst(dds_5_5, blind = TRUE)
vsd_7_8 <- vst(dds_7_8, blind = TRUE)


pca_1_4 <- plotPCA(
  vsd_1_4,
  intgroup = c("Treatment", "Timepoint"),
  returnData = TRUE
)

pca_5_5 <- plotPCA(
  vsd_5_5,
  intgroup = c("Treatment", "Timepoint"),
  returnData = TRUE
)

pca_7_8 <- plotPCA(
  vsd_7_8,
  intgroup = c("Treatment", "Timepoint"),
  returnData = TRUE
)

var_1_4 <- round(100 * attr(pca_1_4, "percentVar"))
var_5_5 <- round(100 * attr(pca_5_5, "percentVar"))
var_7_8 <- round(100 * attr(pca_7_8, "percentVar"))

#########

Fig4_C_14 <- ggplot(pca_1_4, aes(PC1, PC2, color = Treatment, shape = Timepoint)) +
  geom_point(size = 6, alpha = 1) +
  geom_convexhull(aes(fill = Treatment, group = Treatment),alpha = 0.1 ) +
  scale_color_manual(values = amf_colors) +
  xlab(paste0("PC1: ", var_1_4[1], "%")) +
  ylab(paste0("PC2: ", var_1_4[2], "%")) +
  theme_soil()

Fig4_C_55 <- ggplot(pca_5_5, aes(PC1, PC2, color = Treatment, shape = Timepoint)) +
  geom_point(size = 6, alpha = 1) +
  geom_convexhull(aes(fill = Treatment, group = Treatment),alpha = 0.1 ) +
  scale_color_manual(values = amf_colors) +
  xlab(paste0("PC1: ", var_5_5[1], "%")) +
  ylab(paste0("PC2: ", var_5_5[2], "%")) +
  theme_soil()

Fig4_C_78 <- ggplot(pca_7_8, aes(PC1, PC2, color = Treatment, shape = Timepoint)) +
  geom_point(size = 6, alpha = 1) +
  # coord_fixed() +
  geom_convexhull(aes(fill = Treatment, group = Treatment),alpha = 0.1 ) +
  scale_color_manual(values = amf_colors) +
  scale_fill_manual(values = amf_colors) +
  xlab(paste0("PC1: ", var_7_8[1], "%")) +
  ylab(paste0("PC2: ", var_7_8[2], "%")) +
  theme_soil()



ggsave(paste0(workdir, "plots/metaT/Fig4_C_14_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_14, 
       width = 5, height = 4, dpi = 600)


ggsave(paste0(workdir, "plots/metaT/Fig4_C_55_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_55, 
       width = 5, height = 4, dpi = 600)


ggsave(paste0(workdir, "plots/metaT/Fig4_C_78_PCA_metaT_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_C_78, 
       width = 5, height = 4, dpi = 600)

```

## Panel D


```{r}

Fig4_D <- ggplot(combined_data_DV_tr_rhizo, aes(x = Axis.1, y = Axis.2, color = Treat,  fill = Treat, shape = Timepoint)) +
  geom_point(aes(fill = Treat), size = 6, color = "black") +
  geom_convexhull(aes(color = Treat,fill = Treat, group = Treat), alpha = 0.1) +
  scale_color_manual(values = amf_colors) +
  scale_fill_manual(values = amf_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_rhizo$values$Eigenvalues[1] / sum(pcoa_DV_tr_rhizo$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_rhizo$values$Eigenvalues[2] / sum(pcoa_DV_tr_rhizo$values$Eigenvalues) * 100, 2), "%)")) +
  theme_soil()


Fig4_D

ggsave(paste0(workdir, "plots/pcoa/Fig4_D_pcoa_DV_rhizo_AMFtreatment_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_D,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/pcoa/Fig4_D_pcoa_DV_rhizo_AMFtreatment_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_D,
       width = 6, height = 5, dpi = 600)


Fig4_D_legend <- ggplot(combined_data_DV_tr_rhizo, aes(x = Axis.1, y = Axis.2, color = Treat,  fill = Treat, shape = Treat)) +
  geom_point(size = 6, alpha = 1) +
  geom_convexhull(aes(fill = Treat), alpha = 0.1) +
  scale_color_manual(values = amf_colors) +
  scale_fill_manual(values = amf_colors) +
  xlab(paste0("PCoA axis 1 (", round(pcoa_DV_tr_rhizo$values$Eigenvalues[1] / sum(pcoa_DV_tr_rhizo$values$Eigenvalues) * 100, 2), "%)")) +
  ylab(paste0("PCoA axis 2 (", round(pcoa_DV_tr_rhizo$values$Eigenvalues[2] / sum(pcoa_DV_tr_rhizo$values$Eigenvalues) * 100, 2), "%)")) +
  theme_bw()

Fig4_D_legend

ggsave(paste0(workdir, "plots/pcoa/Fig4_D_legend_pcoa_DV_rhizo_AMFtreatment_small_legend", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_D_legend,
       width = 6, height = 5, dpi = 600)

```


## Panel E


```{r}

plant_data_dec23_PCO <- read_csv(paste0(workdir,"plant_data_dec23_plant_chem_only.csv"))


# Filter out T0 and select chemistry columns
plant_chem_filtered <- plant_data_dec23_PCO %>%
  filter(Timepoint != "T0") %>%
  select(`Plant N`:`Plant Na`)  # Replace with actual chemistry columns

# Z-score
plant_chem_scaled <- as.data.frame(scale(plant_chem_filtered))

# PCA
pca_plant_chem <- prcomp(plant_chem_scaled, scale. = TRUE)

# PCA scores
pca_scores_plant_chem <- as.data.frame(pca_plant_chem$x)

# Variance explained
pca_var_plant_chem<- pca_plant_chem$sdev^2
pca_var_perc_plant_chem <- round(100 * pca_var_plant_chem / sum(pca_var_plant_chem), 2)

# Add metadata for plotting
metadata_plant_chem <- plant_data_dec23_PCO %>% filter(Timepoint != "T0")
metadata_plant_chem$Timepoint <- recode(metadata_plant_chem$Timepoint,
                             "T1" = "June",
                             "T2" = "July",
                             "T3" = "August")

# Add metadata
pca_scores_plant_chem <- bind_cols(
  pca_scores_plant_chem,
  metadata_plant_chem
)


pca_scores_plant_chem$Time_month <- factor(
  pca_scores_plant_chem$Timepoint,
  levels = c("June", "July", "August")
)

pca_scores_plant_chem$Site <- case_when(
  pca_scores_plant_chem$`Plot ID` == "1-4 CMT" ~ "1_4",
  pca_scores_plant_chem$`Plot ID` == "5-5 CMT" ~ "5_5",
  pca_scores_plant_chem$`Plot ID` == "7-8 CMT" ~ "7_8",
  TRUE ~ (pca_scores_plant_chem$`Plot ID`)
)


Fig4_E <- ggplot( pca_scores_plant_chem, aes(x = PC1,y = PC2, color = Site, shape = Time_month)) +
  geom_point(aes(fill = Site), size = 6, color = "black") +
  scale_color_manual(values = plot_colors) +
  scale_fill_manual(values = plot_colors) +
  scale_shape_manual(
    name = "Timepoint",
    values = c(22,23,24)
  ) +
  xlab(paste0("PC1 (", pca_var_perc_plant_chem[1], "%)")) +
  ylab(paste0("PC2 (", pca_var_perc_plant_chem[2], "%)")) +
  theme_soil()

Fig4_E


ggsave(paste0(workdir, "plots/plants/Fig4_E_PCA_plants_time_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_E,
       width = 6, height = 5, dpi = 600)

```


```{r}
# Permanova test

library(vegan)

# Prepare the distance matrix using z-scored plant chemistry
plant_chem_dist <- vegdist(plant_chem_scaled, method = "euclidean")

# Prepare metadata for PERMANOVA
metadata_for_permanova <- plant_data_dec23_PCO %>%
  filter(Timepoint != "T0") %>%
  mutate(
    Plot = case_when(
      `Plot ID` == "1-4 CMT" ~ "1_4",
      `Plot ID` == "5-5 CMT" ~ "5_5",
      `Plot ID` == "7-8 CMT" ~ "7_8",
      TRUE ~ `Plot ID`
    )
  )

# Run PERMANOVA
permanova_plant_chem <- adonis2(
  plant_chem_dist ~ Timepoint + Treatment + Plot,
  data = metadata_for_permanova,
  permutations = 999,
  method = "euclidean",  # Already calculated, but specifying for clarity
  by = "terms"
)

# View results
print(permanova_plant_chem)

# Save results to a table
write.csv(as.data.frame(permanova_plant_chem), 
          paste0(workdir, "results/TableS10_PERMANOVA_plant_chemistry_", Sys.Date(), ".csv"))

```

## Panel F


```{r}


# ------------------------------
# 5. Merge DE results with functional annotation
# ------------------------------
DE_annotated <- resOrdered.df %>%
  left_join(emapper_best, by = c("transcript_id" = "query"))

########################################### STATS
# Total genes
n_total <- nrow(DE_annotated)

# Genes passing filter (DEGs)
DE_mod_df <- DE_annotated %>%
  filter(padj < 0.05 & abs(log2FoldChange) > 1) %>%
  mutate(direction = ifelse(log2FoldChange > 0, "Up", "Down"))

n_filtered <- nrow(DE_mod_df)

# Genes filtered out
n_removed <- n_total - n_filtered

# Among filtered genes, how many are annotated? 
# Assuming your annotation column is 'COG_category' or similar
n_annotated <- DE_mod_df %>% 
  filter(!is.na(COG_category) & COG_category != "-") %>% 
  nrow()

# Percent annotated
pct_annotated <- n_annotated / n_filtered * 100

# Print summary
cat("Total genes:", n_total, "\n")
cat("Genes passing filter (DEGs):", n_filtered, "\n")
cat("Genes filtered out:", n_removed, "\n")
cat("Annotated DEGs:", n_annotated, "(", round(pct_annotated, 2), "%)\n")


#########################################

# Summarize average log2FC by COG and direction

DE_mod_df <- DE_mod_df %>%
  mutate(COG_category = ifelse(COG_category == "-", NA, COG_category))

# --- COG DESCRIPTION LOOKUP ---
COG_desc_map <- c(
  # INFORMATION STORAGE AND PROCESSING
  "J" = "Translation, ribosomal structure, and biogenesis",
  "A" = "RNA processing and modification",
  "K" = "Transcription",
  "L" = "Replication, recombination, and repair",
  "B" = "Chromatin structure and dynamics",
  
  # CELLULAR PROCESSES AND SIGNALING
  "D" = "Cell cycle control, cell division, chromosome partitioning",
  "Y" = "Nuclear structure",
  "V" = "Defense mechanisms",
  "T" = "Signal transduction mechanisms",
  "M" = "Cell wall/membrane/envelope biogenesis",
  "N" = "Cell motility",
  "Z" = "Cytoskeleton",
  "W" = "Extracellular structures",
  "U" = "Intracellular trafficking, secretion, and vesicular transport",
  "O" = "Posttranslational modification, protein turnover, chaperones",
  
  # METABOLISM
  "C" = "Energy production and conversion",
  "G" = "Carbohydrate transport and metabolism",
  "E" = "Amino acid transport and metabolism",
  "F" = "Nucleotide transport and metabolism",
  "H" = "Coenzyme transport and metabolism",
  "I" = "Lipid transport and metabolism",
  "P" = "Inorganic ion transport and metabolism",
  "Q" = "Secondary metabolite biosynthesis, transport, and catabolism",
  
  # POORLY CHARACTERIZED
  "R" = "General function prediction only",
  "S" = "Function unknown",
  
  # Unannotated
  "Unannotated" = "Unannotated"
)

# 1. Combine Unannotated + "S" (Function unknown)
DE_mod_df2 <- DE_mod_df %>%
  mutate(
    COG = COG_category,
    COG = ifelse(COG %in% c("-", NA, "S"), "Function unknown", COG)
  )

# 2. Count genes per COG × direction
summary_COG <- DE_mod_df2 %>%
  group_by(COG, direction) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
  
  # 3. Make downregulated negative
  mutate(n_genes_signed = ifelse(direction == "Down",
                                 -n_genes,
                                 n_genes))

# 4. Add descriptions
summary_COG <- summary_COG %>%
  mutate(
    COG_desc = case_when(
      COG == "Function unknown" ~ "Function unknown",
      COG %in% names(COG_desc_map) ~ COG_desc_map[COG],
      TRUE ~ "Mixed functional category"
    )
  )


direction_colors <- c(
  "Down"   = "#8C8C8C",
  "Up" = "#2F3E5C"   # deep navy
)

summary_COG_mod <- summary_COG %>%  filter(COG_desc != 'Mixed functional category' & COG_desc != 'Function unknown')

dim(summary_COG)
dim(summary_COG_mod)

Fig4_F <- ggplot(summary_COG_mod,
       aes(x = reorder(COG_desc, n_genes_signed),
           y = n_genes_signed,
           fill = direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = direction_colors) +
  coord_flip() +
  xlab("COG Category") +
  ylab("Number of Differentially Expressed Genes") +
  theme_soil()

Fig4_F <- ggplot(summary_COG_mod,
       aes(x = reorder(n_genes_signed),
           y = n_genes_signed,
           fill = direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = direction_colors) +
  coord_flip() +
  xlab("COG Category") +
  ylab("Number of Differentially Expressed Genes") +
  theme_soil()

Fig4_F

ggsave(paste0(workdir, "plots/metaT/Fig4_F_DE_up_down_regulated_genes_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_F,
       width = 7, height = 6, dpi = 600)

ggsave(paste0(workdir, "plots/metaT/Fig4_F_DE_up_down_regulated_genes_2_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_F,
       width = 12, height = 6, dpi = 600)

########### Re ordered by the numebr of downregulated genes:

cog_order <- summary_COG_mod %>%
  filter(direction == "Down") %>%
  arrange(n_genes_signed) %>%   # more down = more negative
  pull(COG_desc)

summary_COG_mod$COG_desc <- factor(
  summary_COG_mod$COG_desc,
  levels = cog_order
)

# Recode the names to shorter versions

cog_short <- c(
  "RNA processing and modification"                               = "A: RNA processing",
  "Energy production and conversion"                              = "C: Energy",
  "Cell cycle control, cell division, chromosome partitioning"    = "D: Cell cycle",
  "Amino acid transport and metabolism"                           = "E: Amino acids",
  "Nucleotide transport and metabolism"                           = "F: Nucleotides",
  "Carbohydrate transport and metabolism"                         = "G: Carbohydrates",
  "Coenzyme transport and metabolism"                             = "H: Coenzymes",
  "Lipid transport and metabolism"                                = "I: Lipids",
  "Translation, ribosomal structure, and biogenesis"              = "J: Translation",
  "Transcription"                                                 = "K: Transcription",
  "Replication, recombination, and repair"                        = "L: Replication and repair",
  "Cell wall/membrane/envelope biogenesis"                        = "M: Cell envelope",
  "Posttranslational modification, protein turnover, chaperones" = "O: Post-translation",
  "Inorganic ion transport and metabolism"                        = "P: Inorganic ions",
  "Secondary metabolite biosynthesis, transport, and catabolism" = "Q: Secondary metabolism",
  "Signal transduction mechanisms"                                = "T: Signal transduction",
  "Intracellular trafficking, secretion, and vesicular transport"= "U: Vesicular transport",
  "Defense mechanisms"                                            = "V: Defense"
)

summary_COG_mod <- summary_COG_mod %>%
  mutate(COG_short = recode(COG_desc, !!!cog_short))


Fig4_F_2 <- ggplot(summary_COG_mod,
       aes(x = COG_short,
           y = n_genes_signed,
           fill = direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = direction_colors) +
  coord_flip() +
  xlab("COG Category") +
  ylab("Number of Differentially Expressed Genes") +
  theme_soil()

Fig4_F_2

# Labels aligned
Fig4_F_2 <- ggplot(summary_COG_mod,
       aes(x = COG_short,
           y = n_genes_signed,
           fill = direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = direction_colors) +
  coord_flip() +
  xlab("COG Category") +
  ylab("Number of Differentially Expressed Genes") +
  theme_soil() +
  theme(
    axis.text.y = element_text(hjust = 0)  # ← THIS LINE
  )

Fig4_F_2


ggsave(paste0(workdir, "plots/metaT/Fig4_F_2_DE_up_down_regulated_genes_small_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Fig4_F_2,
       width = 8, height = 6, dpi = 600)

```


```{r}

Fig4_F <- ggplot(summary_COG, aes(x = reorder(COG_desc, n_genes_signed), y = n_genes_signed, fill = direction)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = direction_colors) +
  coord_flip() +
  theme_soil(base_size = 14)+
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 8)  # make the category labels smaller
  ) +
  ylab("Number of DE Genes") +
  xlab("COG Category")


```


## Figure 4 combined

```{r}
# Combine with patchwork


panel_E <- (Fig4_C_T1 | Fig4_C_T2 | Fig4_C_T3) +   # combine 3 plots horizontally
  plot_layout(guides = "collect") & 
  theme(plot.margin = margin(1,1,1,1))


bottom_row_Fig4 <- (panel_E | Fig4_F_2) + 
  plot_layout(widths = c(1,1,1,3), guides = "collect") &
  theme(plot.margin = margin(2,2,2,2))


# Bottom row: C and D side by side
top_rows_Fig4 <-   (Fig4_A | Fig4_D) /
  (Fig4_B | Fig4_E) &
  theme(plot.margin = margin(2,2,2,2))

# Combine top and bottom rows
Figure_4_combined <- top_rows_Fig4 / bottom_row_Fig4 +
  #plot_layout(heights = c(1,1)) +  # same height for top & bottom
  plot_annotation(tag_levels = 'A')

Figure_4_combined

ggsave(paste0(workdir, "plots/Manuscript_figures/Figure_4_combined_", Sys.Date(), ".pdf"), 
       device = "pdf",
       plot = Figure_4_combined, 
       width = 12, height = 16, dpi = 600)



```

